---
title: "first calcs"
format: docx
editor: visual
---

```{r load data}

library(tidyverse)
income = read.csv("P:/CH-Ranking/Research/Disparity Measurement/Census Tract Disparity Measurement/Duplication_GL/v153_disparity_income_GL.csv")


race = read.csv("P:/CH-Ranking/Research/Disparity Measurement/Census Tract Disparity Measurement/Duplication_GL/v153_disparity_race_GL.csv")

tract = read.csv("P:/CH-Ranking/Research/Disparity Measurement/Census Tract Disparity Measurement/Duplication_GL/v153_disparity_tract_GL.csv")

#mean_wt and mean_th_wt are equivalent 
#num_hh and num_hh_theil_counted are equivalent 

```

Theils index is decomposable: sum of between group disparity (sum(inds pop share and weighted avg of within group disparity))

From p50 of Harper and Lynch:

For individual-level data, total disparity in mortality (let's say) measured by Theil’s index can be written as T = N Σi = 1 Σ ln (ri ) pi sj ri 1n pj sj ( \[19\] where pi is an individual’s population share (which in the case of individual data will be 1/n, so that pi = 1) and ri is the ratio of the individual’s age of mortality to the population average age of mortality (i.e., ri = yi / Y ( ). (in this case let's assume that the total population is a single county)

When the population of individuals is arranged into J groups (say census tracts), Theil showed that equation \[19\] is the exact sum of two parts: between-census tract disparity and a weighted average of within-census tract disparity: T = J Σj = 1 ln (rj ) + pj rj J Σj = 1 pj rj Tj \[20\] where Tj is the disparity in BMI within group j. The within-group component (the second term on the right side of equation \[20\] is weighted by, in this case, group j’s share of the total (county-level) age at mortality, pj x rj = sj because (where sj is the share of total BMI) when the denominator for rj is the mean age at mortality for the total (county) population. More importantly, the above decomposition also makes it clear that it is possible to calculate between-census tract disparity in mortality age—the primary quantity of interest with respect to social disparities in health—in the absence of data on each individual.

BIG TAKE AWAY: "Although this measure has attractive qualities, the betweengroup/within-group decomposition requires continuous outcome data measurable in individuals"

BUT we could possibly consider CTs/race groups/income groups as "individuals", counties as groups, and states as "total"

which means that we could compare the between county/ within county decomposition across income, race, and tracts.

## Multiple regression analysis

dependent var: CT BGV

independent var: income BGV and race BGV

The coeffs of this model will tell me the strength and direction of each relationship

R2 will tell me how much of the total variation in CT disparities is explained by income and race

```{r}
race = race %>% rename(sd_bgv_wt_race = sd_bgv_wt) 
income = income %>% rename(sd_bgv_wt_income = sd_bgv_wt)
tract = tract %>% rename(sd_bgv_wt_tract = sd_bgv_wt)

all = income %>% 
  left_join(race, by = "fipscode") %>%
  left_join(tract, by = "fipscode")

model <- lm(sd_bgv_wt_tract ~ sd_bgv_wt_income + sd_bgv_wt_race, data = all)

```

draft results para: The results of the multiple regression analysis indicate that both **income disparities** and **race disparities** significantly contribute to explaining **census tract disparities** within counties, but **income disparities** appear to have a much stronger relationship. Specifically, for each one-unit increase in the income-based BGV value, census tract disparities are expected to increase by 0.576 units, with a highly significant p-value (\< 2e-16), suggesting that income disparities play a dominant role in shaping disparities at the tract level. In contrast, for each one-unit increase in the race-based BGV value, census tract disparities increase by 0.083 units, with a significant p-value (0.0035), though the effect size is smaller than that of income. These findings suggest that while both race and income disparities are important, **income disparities** have a stronger and more statistically significant influence on the disparities across census tracts within counties. The model’s R-squared value indicates that while the predictors explain only a small portion (11.07%) of the variance in census tract disparities, the relationship between both income and race disparities and census tract disparities is statistically significant.

## Partial correlation

```{r}

library(ppcor)

# Calculate partial correlation controlling for income
partial_corr_income <- pcor.test(all$sd_bgv_wt_tract, all$sd_bgv_wt_race, all$sd_bgv_wt_income)

# Calculate partial correlation controlling for race
partial_corr_race <- pcor.test(all$sd_bgv_wt_tract, all$sd_bgv_wt_income, all$sd_bgv_wt_race)

# Print partial correlations
partial_corr_income
partial_corr_race

```

Draft results para:

The results of the partial correlation analysis reveal the strength of the relationship between **census tract disparities** and each of the predictors (**race disparities** and **income disparities**), while controlling for the influence of the other. The partial correlation between **census tract disparities** and **race disparities** is **0.299**, with a p-value of **\< 2e-16**, indicating a **moderate positive** relationship between these two variables that is highly statistically significant. This suggests that, after accounting for income disparities, race disparities have a notable influence on the disparities observed at the census tract level. In contrast, the partial correlation between **census tract disparities** and **income disparities** is much weaker, with an estimate of **0.052** and a p-value of **0.0035**, indicating a **small positive relationship** that is still statistically significant but much less influential than race disparities. This suggests that income disparities, while significant, have a relatively minor effect on **census tract disparities** compared to race disparities when controlling for the other variable.

NEED TO CHECK THIS; CONFUSED WHY MODEL COEFFS AND PCOR RESULTS ARE DISCORDANT

## ABSTRACT draft

We examine patterns in county-level disparities in homeownership, as measured by between-group variance, with the goal of understanding what **census tract-level disparities** are actually capturing—whether they are primarily reflecting **race differences**, **income differences**, or a combination of both. We compare **income-based disparities** (defined using 22 income groups), **race/ethnicity-based disparities** (defined using 8 different racial/ethnic identity groups), and **geographical disparities** (defined using census tracts). Using multiple regression and partial correlation analysis, the results show that **race disparities** are more strongly associated with census-tract level disparities than **income disparities**. Specifically, the partial correlation between race disparities and census-tract disparities is moderate (0.299), while the relationship with income disparities is weak (0.052). Both relationships are statistically significant, but the stronger influence of race highlights the importance of considering racial factors in understanding geographical disparities in homeownership. These findings suggest that **race** may play a more prominent role than income in shaping disparities in homeownership at the county level.

```{r}
# Load necessary libraries
library(lme4)  # For mixed-effects models
library(dplyr) # For data manipulation



# Fit a two-level linear mixed-effects model
# The response variable is 'fv_servings', with individual-level predictors
# and tract-level predictors, and a random intercept for each tract.

model <- lmer(fv_servings ~ race_ethnicity + nativity + duration_in_us + 
               language_acculturation + emotional_support + instrumental_support + 
               socioeconomic_position + 
               tract_immigrant_composition + tract_racial_composition + 
               tract_socioeconomic_composition + 
               (1 | tract_id), data = data)

# Summary of the model to assess fixed and random effects
summary(model)

# You can extract the coefficients (fixed effects) and random effects if needed
fixef(model)  # Fixed effects
ranef(model)  # Random effects (tract-level variation)

# For model diagnostics, you can also check residuals
residuals(model)  # Check model residuals

```
