---
title: "Rough draft"
#format: pdf
execute:
  echo: false      # This hides the code from being printed
  warning: false   # This suppresses warnings
  message: false   # This suppresses messages
  error: false
  results: 'markup' # This ensures that only results are displayed
output:
  html_document: 
  self_contained: true 
---

```{r get data into a usable form }
library(tidyverse)
# Load each file and assign to a new variable name
assign("tract", get(load("calcd_data_how/tract.RData")))
assign("tract_cis", get(load("calcd_data_how/tract_cis.RData")))
assign("race", get(load("calcd_data_how/race.RData")))
assign("race_cis", get(load("calcd_data_how/race_cis.RData")))
assign("income", get(load("calcd_data_how/income.RData")))
assign("income_cis", get(load("calcd_data_how/income_cis.RData")))

# Function to rename columns except the first three
rename_columns_cis <- function(df, dataset_name) {
  df %>%
    rename_with(~ ifelse(. %in% c("statecode", "countycode", "fipscode", "GEOID"), ., paste0(dataset_name, "_", .)))
  #, -c(statecode, countycode, fipscode, GEOID))
}

# Rename columns for tract_cis, race_cis, and income_cis
tract_cis <- rename_columns_cis(tract_cis, "tract_cis")
race_cis <- rename_columns_cis(race_cis, "race_cis")
income_cis <- rename_columns_cis(income_cis, "income_cis")
tract <- rename_columns_cis(tract, "tract")
race <- rename_columns_cis(race, "race")
income <- rename_columns_cis(income, "income")

# Assuming the objects are named tract, tract_cis, race, race_cis, income, and income_cis
all <- tract %>%
  left_join(tract_cis, by = c("statecode", "countycode")) %>%
  left_join(race, by = c("statecode", "countycode")) %>%
  left_join(race_cis, by = c("statecode", "countycode")) %>%
  left_join(income, by = c("statecode", "countycode")) %>%
  left_join(income_cis, by = c("statecode", "countycode"))

```

```{r get the mapping function ready}
library(tigris)
library(ggplot2)
library(dplyr)
library(sf)

# Load county shapefile
#counties <- counties(cb = TRUE, resolution = '20m') %>%  st_as_sf() %>%  rename(statecode = STATEFP, countycode = COUNTYFP) 
#save(counties, file = "inputs/counties_shp.RData")

# Merge data with county shapefile
load("inputs/counties_shp.RData")

map_data <- left_join(counties, all, by = c("statecode", "countycode"))

# Select columns that contain "sd_bgv_wt" in their name
sd_bgv_wt_columns <- map_data %>% 
  select(contains("sd_bgv_wt")) %>% st_drop_geometry()

# Calculate global minimum and maximum of the selected columns
#global_min <- min(sd_bgv_wt_columns, na.rm = TRUE)
#global_max <- max(sd_bgv_wt_columns, na.rm = TRUE)

# Set specific columns to NA if they are 0
map_data <- map_data %>%
  mutate(across(c(tract_cis_sd_bgv_wt, race_cis_sd_bgv_wt, income_cis_sd_bgv_wt), 
                ~if_else(. == 0, NA_real_, .)))


# Create map plotting function
plot_map <- function(column, title, legend = TRUE) {
  ggplot(map_data) +
    geom_sf(aes_string(fill = column), color = NA) +
    scale_fill_viridis_c(option = "plasma", na.value = "grey80", limits = c(min(map_data[[column]], na.rm = TRUE), max(map_data[[column]], na.rm = TRUE))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.margin = margin(5, 5, 5, 5),
      panel.grid = element_blank(),
      legend.position = if (legend) "right" else "none"
    ) +
    coord_sf(crs = 5070) +  # Standard Albers Equal Area projection
    labs(fill = title, title = title)
}




```

# An examination of disparities in homeownership across three domains: race/ethnicity, income, and geography.

#### **1. Introduction**

**Purpose of Study** 

-   Compare and quantify within-county disparities in homeownership across three domains (income, race/ethnicity, and geography)

-   Compare and quantify the relationships between within-county disparities in homeownership across domains and county-level measures of health (premature mortality, f/p health, etc)

**Research Questions and Hypotheses** 

-   What are the magnitudes of disparity within counties for each domain?

-   Are there state-level differences in disparity magnitudes and correlations? What conclusions can we draw based on these state-level differences? 

-   What are the relationships between disparity within counties across each domain and county-level health?

-   Something about missingness and data availability

#### 2. Literature review: 

-   Why are disparities in homeownership important? Connection to power, wealth, structure, etc

    -   historical context

    -   future context

-   What does current lit say about relative size of income versus racial disparities in home ownership

-   What does current lit say about the importance of measuring disparities across domains

-   What does current lit say about the correlations between disparity and health?

### Methods 

We begin by calculating homeownership rates using data from the 2018-2022 American Community Survey. The numerator is the total number of owner-occupied housing units and the denominator is the total number of occupied housing units. We calculate homeownership rates at the county-level for all US counties, at the census tract level, at the race-level within counties, and at the income-level within counties. We set homeownership rates to missing when the rate is unreliable, as indicated by a 95% confidence interval that falls outside of possible values (ie inclusive of rates that are less than 0 or greater than 1).

Next we calculate within county disparities separately across race/ethnicity groups, across income groups, and across census tracts. We calculate within-county disparities using the standard deviation of population-weighted group means

$$ insert equation here $$ 

We chose to use the standard deviation of population-weighted group means because it produced normally distributed values and we found it to be robust to outliers. We also wanted a population-weighted group mean so that small populations would not overly skew the disparity metric.

### Results 

table here of just county-level homeownership rates

map of county-level homeownership rates (to show missingness patterns)

## Maps with unstable estimates removed

Median homeownership rates of counties with non-missing data by income, census tracts, and race/ethnicity were 73.0%, 71.7%, and 71.4%, respectively, with most missingness occuring across the western plains and central southesast. Within-county disparities were largest by income (11.9%), followed by race/ethnicity (9.5%) and census tracts (9.4%). The coasts (California and New England) appear to have high tract and income disparities but possibly lower race disparities.

```{r}
# Plot each map
plot_tract <- plot_map("tract_cis_sd_bgv_wt", "SD across Tracts", legend = FALSE)
plot_income <- plot_map("income_cis_sd_bgv_wt", "SD across Income", legend = FALSE)
plot_race_wlegend <- plot_map("race_cis_sd_bgv_wt", "SD across Race/Ethnicities")

plot_tract
plot_income
plot_race_wlegend
```

## Histograms of SD for Race, Income, Tract

The majority of missingness occurs in rural counties where there are fewer

```{r histograms for urb cats}


urbcodes = readxl::read_xlsx("inputs/Ruralurbancontinuumcodes2023.xlsx")
urbcodes <- urbcodes %>% mutate(FIPS = as.numeric(FIPS))  # Ensure numeric for join
all = all %>% mutate(fipscode = as.numeric(fipscode.x))
all_urb <- left_join(all, urbcodes, by = c("fipscode" = "FIPS"))


# Set specific columns to NA if they are 0
hist_data <- all_urb %>%
  mutate(across(c(tract_cis_sd_bgv_wt, race_cis_sd_bgv_wt, income_cis_sd_bgv_wt), 
                ~if_else(. == 0, NA_real_, .)))


tworucc = hist_data %>% group_by(RUCC_2023) %>% 
  mutate(RUCC_group = ifelse(RUCC_2023 %in% c(1,2,3), "Urban", "Rural"))


library(grid)
# Create histograms for specified RUCC_2023 code with additional options for cis and BorT
plot_histograms_by_rucc <- function(rucc_group) {
  
  # Filter data for the specified RUCC code
  filtered_data <- tworucc %>% filter(RUCC_group == rucc_group)

 
  # Create histograms for each category
  hist1 <- ggplot(filtered_data, aes(x = race_cis_sd_bgv_wt)) +
    geom_histogram(bins = 30, fill = "#1f77b4", alpha = 0.7) +
    theme_minimal(base_size = 14) + 
    labs(x = "Race", y = "Count")

  hist2 <- ggplot(filtered_data, aes(x = tract_cis_sd_bgv_wt)) +
    geom_histogram(bins = 30, fill = "#ff7f0e", alpha = 0.7) +
    theme_minimal(base_size = 14) +
    labs(x = "Tract", y = "Count")

  hist3 <- ggplot(filtered_data, aes(x = income_cis_sd_bgv_wt)) +
    geom_histogram(bins = 30, fill = "#2ca02c", alpha = 0.7) +
    theme_minimal(base_size = 14) +
    labs(x = "Income", y = "Count")
  
  # Define dynamic title
  main_title <- paste0("SD: ", rucc_group)

  

  # Arrange the plots in a grid with a title
  gridExtra::grid.arrange(
    grobs = list(
      grid::textGrob(main_title, gp = gpar(fontsize = 16, fontface = "bold")),  # Title
      hist1, hist2, hist3
    ))
  
}

plot_histograms_by_rucc(rucc_group = "Rural")
plot_histograms_by_rucc(rucc_group = "Urban")

```

percent missing

```{r this doesnt work yet - goal is to add percent missing to the hist above}




# Calculate percent missing for each group
missing_summary <- tworucc %>%
  group_by(RUCC_2023) %>%
  summarise(across(where(is.numeric), ~mean(is.na(.)) * 100, .names = "missing_{.col}")) %>%
  pivot_longer(cols = starts_with("missing_"), 
               names_to = "variable", 
               values_to = "percent_missing") %>%
  mutate(variable = gsub("missing_", "", variable))

# Plot
ggplot(missing_summary, aes(x = variable, y = percent_missing, fill = rurality)) +
  geom_col(position = "dodge") +
  labs(title = "Percent Missing by Rurality",
       x = "Variable",
       y = "Percent Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Correlations

Tract disparities and income disparities are more strongly correlated than tract disparities and race disparities. Income and race disparities are /slightly/ more correlated than tract and race disparities. The correlations below are calculated using Pearson's correlation coefficient and assume Normality and linearity.

```{r correlations}



library(corrplot)
library(ggplot2)
library(pheatmap)

# Assuming the necessary variables are in 'map_data'
# Select the relevant columns for correlation analysis
cor_data <- hist_data %>%
  select(
    tract_cis_sd_bgv_wt, 
    income_cis_sd_bgv_wt, 
    race_cis_sd_bgv_wt, 
  )

# Calculate correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs", method = "pearson")

# Visualize the correlation matrix using corrplot
corrplot(cor_matrix, method = "color", type = "upper", 
         col = colorRampPalette(c("pink", "white", "lightblue"))(200),
         title = "Correlation Heatmap", 
         mar = c(0,0,1,0))

# Alternatively, you can use pheatmap for a more interactive approach
pheatmap(cor_matrix, 
         display_numbers = TRUE, 
         color = colorRampPalette(c("pink", "white", "lightblue"))(100),
         main = "Correlation Heatmap", 
         fontsize = 12)

```

The strongest correlation occurs between tract and income, weakest between tract and race. All correlations are significant. The correlations below are calculated using Spearman's correlation test which is robust to nonlinear relationships.

```{r}
# Example: Calculate Spearman's correlation
race_inc <- cor.test(hist_data$race_cis_sd_bgv_wt, hist_data$income_cis_sd_bgv_wt, method = "spearman")

tract_inc <- cor.test(hist_data$tract_cis_sd_bgv_wt, hist_data$income_cis_sd_bgv_wt, method = "spearman")

tract_race <- cor.test(hist_data$tract_cis_sd_bgv_wt, hist_data$race_cis_sd_bgv_wt, method = "spearman")

race_inc
tract_inc
tract_race

```

## State-level rhos

```{r state level rho calcs }
library(dplyr)
library(purrr)

# Updated function to safely calculate Spearman's rho
get_spearman_rho <- function(x, y) {
  if (length(na.omit(x)) < 2 || length(na.omit(y)) < 2) {
    return(NA)  # Not enough data to compute correlation
  }
  cor.test(x, y, method = "spearman")$estimate
}

# Calculate Spearman's rho per state
rho_by_state <- hist_data %>%
  group_by(State) %>%
  summarise(
    race_inc_rho = get_spearman_rho(race_cis_sd_bgv_wt, income_cis_sd_bgv_wt),
    tract_inc_rho = get_spearman_rho(tract_cis_sd_bgv_wt, income_cis_sd_bgv_wt),
    tract_race_rho = get_spearman_rho(tract_cis_sd_bgv_wt, race_cis_sd_bgv_wt)
  )

```

```{r state level map function }
library(ggplot2)
library(maps)

# Load US states map
states_map <- map_data("state")

# Convert statecode to lowercase full names (you may need a lookup table here)
# This assumes statecode is a standard two-letter code
library(tibble)

state_lookup <- tibble::tibble(
  State = state.abb,
  statename = tolower(state.name)
)

# Join rho data with full state names
rho_by_state_map <- rho_by_state %>%
  left_join(state_lookup, by = "State") %>%
  filter(!is.na(statename)) # Remove states not matched

```

```{r maps themselves }
# Merge with map data
plot_data <- states_map %>%
  left_join(rho_by_state_map, by = c("region" = "statename"))

# Define a plotting function
plot_rho_map <- function(data, fill_col, title) {
  ggplot(data, aes(x = long, y = lat, group = group, fill = .data[[fill_col]])) +
    geom_polygon(color = "white") +
    coord_fixed(1.3) +
    scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
    labs(title = title, fill = "Spearman's ρ") +
    theme_minimal()
}

# Create each map
map_race_inc <- plot_rho_map(plot_data, "race_inc_rho", "Race vs. Income Correlation by State")
map_tract_inc <- plot_rho_map(plot_data, "tract_inc_rho", "Tract vs. Income Correlation by State")
map_tract_race <- plot_rho_map(plot_data, "tract_race_rho", "Tract vs. Race Correlation by State")

# Display the maps
print(map_race_inc)
print(map_tract_inc)
print(map_tract_race)

```

## State-level SD

```{r Calculate the median for each state}

median_by_state <- hist_data %>%
  group_by(State) %>%
  summarise(
    median_tract = median(tract_cis_sd_bgv_wt, na.rm = TRUE),
    median_race = median(race_cis_sd_bgv_wt, na.rm = TRUE),
    median_income = median(income_cis_sd_bgv_wt, na.rm = TRUE),
    .groups = "drop"
  )




# Merge the median values with state names
median_by_state_map <- median_by_state %>%
  left_join(state_lookup, by = "State") %>%
  filter(!is.na(statename))  # Remove unmatched states


```

```{r make the map and print top 3 below}
library(ggplot2)
library(dplyr)

# Merge the state-level median values with the map data
plot_data <- states_map %>%
  left_join(median_by_state_map, by = c("region" = "statename"))

# Function to get the states with the highest median values
get_top_states <- function(data, fill_col, top_n = 3) {
  data %>%
    group_by(region) %>%
    summarise(top_value = max(.data[[fill_col]], na.rm = TRUE)) %>%
    arrange(desc(top_value)) %>%
    head(top_n) %>%
    pull(region)
}

# Function to plot the map and annotate the top states
plot_median_map <- function(data, fill_col, title, top_states) {
  ggplot(data, aes(x = long, y = lat, group = group, fill = .data[[fill_col]])) +
    geom_polygon(color = "white") +
    coord_fixed(1.3) +
    scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
    labs(title = title, fill = "Median Value") +
    theme_minimal() +
    annotate("text", x = -125, y = 25, label = paste("Top States: ", paste(top_states, collapse = ", ")), 
             size = 5, color = "black", hjust = 0, vjust = 0)
}

# Get the top 5 states for each variable
top_states_tract <- get_top_states(plot_data, "median_tract")
top_states_race <- get_top_states(plot_data, "median_race")
top_states_income <- get_top_states(plot_data, "median_income")

# Create the maps for each variable
map_median_tract <- plot_median_map(plot_data, "median_tract", "Median Tract Disparity by State", top_states_tract)
map_median_race <- plot_median_map(plot_data, "median_race", "Median Race Disparity by State", top_states_race)
map_median_income <- plot_median_map(plot_data, "median_income", "Median Income Disparity by State", top_states_income)

# Display the maps
print(map_median_tract)
print(map_median_race)
print(map_median_income)

```

## States where tract disparities are greater than income disparities

```{r}
median_by_state %>% filter(median_income < median_tract)



```

## States where race disparities are greater than income disparities

```{r}
median_by_state %>% filter(median_income < median_race)
```

## Missingness

```{r venn diagram}

library(ggvenn)


# Step 2: Create subsets for the Venn diagram based on missing values
venn_data <- list(
  race = hist_data %>% filter(is.na(race_cis_sd_bgv_wt)) %>% pull(fipscode),
  tract = hist_data %>% filter(is.na(tract_cis_sd_bgv_wt)) %>% pull(fipscode),
  income = hist_data %>% filter(is.na(income_cis_sd_bgv_wt)) %>% pull(fipscode)
)

# Calculate total number of missing entries (only consider missing data)
total_missing <- length(unique(c(venn_data$race, venn_data$tract, venn_data$income)))



# Create Venn diagram using ggvenn
ggvenn(venn_data, 
       show_percentage = TRUE, 
       fill_color = c("lightblue", "lightgreen", "lightcoral"),
       stroke_color = "black")
      # title = "Venn Diagram of Missing Values for Race, Tract, and Income SD")

```

```{r table of missingness}
# Load the gt package
library(gt)

# Create the missing_counts dataframe
missing_counts <- data.frame(
  category = c("Race", "Tract", "Income"),
  count = c(length(venn_data$race), length(venn_data$tract), length(venn_data$income))
)

# Calculate percent missing for each category
missing_counts$percent_missing <- (missing_counts$count / nrow(hist_data)) * 100

# Use gt to create a simple, pretty table
missing_counts %>%
  gt() %>%
  tab_header(
    title = "Missing Data by Category"
  ) %>%
  cols_label(
    category = "Category",
    count = "Count",
    percent_missing = "Percent Missing"
  ) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black", weight = px(2)),
    locations = cells_body()
  ) %>%
  tab_options(
    table.width = pct(80),
    column_labels.font.size = 14,
    table.border.top.color = "black",
    table.border.bottom.color = "black"
  )

```

### Summary stats with two RUCC groups

Split is here: <https://www.ers.usda.gov/data-products/rural-urban-continuum-codes/documentation#:~:text=For%20Rural%2DUrban%20Continuum%20Codes,below%20250%2C000%20(code%203).>

```{r latest summary table}



# Load necessary libraries
library(dplyr)
library(gt)

# Assuming 'tworucc' is your dataset, and you have a column named 'RUCC_group'
# Calculate missing and available percentages, along with median, min, max for each variable

tworucc_summary <- tworucc %>%
  mutate(
    # Calculate missing data for each variable
    race_missing = sum(is.na(race_cis_sd_bgv_wt)),
    income_missing = sum(is.na(income_cis_sd_bgv_wt)),
    tract_missing = sum(is.na(tract_cis_sd_bgv_wt)),
    total_count = n(),
    
    # Calculate percent data available
    race_available_percent = (1 - race_missing / total_count) * 100,
    income_available_percent = (1 - income_missing / total_count) * 100,
    tract_available_percent = (1 - tract_missing / total_count) * 100
  ) %>%
  group_by(RUCC_group) %>%
  summarise(
    # Calculate missing counts
    race_missing = sum(is.na(race_cis_sd_bgv_wt)),
    income_missing = sum(is.na(income_cis_sd_bgv_wt)),
    tract_missing = sum(is.na(tract_cis_sd_bgv_wt)),
    
    # Calculate available percentages
    race_available_percent = (1 - race_missing / n()) * 100,
    income_available_percent = (1 - income_missing / n()) * 100,
    tract_available_percent = (1 - tract_missing / n()) * 100,
    
    # Median, min, and max for race, income, and tract (combined)
    race_stats = paste(
      round(median(race_cis_sd_bgv_wt, na.rm = TRUE), 4),
      "(", round(min(race_cis_sd_bgv_wt, na.rm = TRUE), 4),
      ",", round(max(race_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
    ),
    
    income_stats = paste(
      round(median(income_cis_sd_bgv_wt, na.rm = TRUE), 4),
      "(", round(min(income_cis_sd_bgv_wt, na.rm = TRUE), 4),
      ",", round(max(income_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
    ),
    
    tract_stats = paste(
      round(median(tract_cis_sd_bgv_wt, na.rm = TRUE), 4),
      "(", round(min(tract_cis_sd_bgv_wt, na.rm = TRUE), 4),
      ",", round(max(tract_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
    ),
    
    .groups = "drop"
  ) %>%
  bind_rows(
    # Calculate overall summary statistics (combined)
    tibble(
      RUCC_group = "Overall",
      race_missing = sum(is.na(tworucc$race_cis_sd_bgv_wt)),
      income_missing = sum(is.na(tworucc$income_cis_sd_bgv_wt)),
      tract_missing = sum(is.na(tworucc$tract_cis_sd_bgv_wt)),
      
      race_available_percent = (1 - sum(is.na(tworucc$race_cis_sd_bgv_wt)) / nrow(tworucc)) * 100,
      income_available_percent = (1 - sum(is.na(tworucc$income_cis_sd_bgv_wt)) / nrow(tworucc)) * 100,
      tract_available_percent = (1 - sum(is.na(tworucc$tract_cis_sd_bgv_wt)) / nrow(tworucc)) * 100,
      
      race_stats = paste(
        round(median(tworucc$race_cis_sd_bgv_wt, na.rm = TRUE), 4),
        "(", round(min(tworucc$race_cis_sd_bgv_wt, na.rm = TRUE), 4),
        ",", round(max(tworucc$race_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
      ),
      
      income_stats = paste(
        round(median(tworucc$income_cis_sd_bgv_wt, na.rm = TRUE), 4),
        "(", round(min(tworucc$income_cis_sd_bgv_wt, na.rm = TRUE), 4),
        ",", round(max(tworucc$income_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
      ),
      
      tract_stats = paste(
        round(median(tworucc$tract_cis_sd_bgv_wt, na.rm = TRUE), 4),
        "(", round(min(tworucc$tract_cis_sd_bgv_wt, na.rm = TRUE), 4),
        ",", round(max(tworucc$tract_cis_sd_bgv_wt, na.rm = TRUE), 4), ")"
      )
    )
  )

# Create the table using gt
tworucc_summary %>%
  gt() %>%
  tab_header(
    title = "Summary of Homeownership Rates by RUCC Group"
  ) %>%
  cols_label(
    RUCC_group = "RUCC Group",
    race_available_percent = "Race Data Available (%)",
    income_available_percent = "Income Data Available (%)",
    tract_available_percent = "Tract Data Available (%)",
    race_stats = "Race (Median, Min, Max)",
    income_stats = "Income (Median, Min, Max)",
    tract_stats = "Tract (Median, Min, Max)"
  ) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black", weight = px(2)),
    locations = cells_body()
  ) %>%
  tab_options(
    table.width = pct(90),
    column_labels.font.size = 14,
    table.border.top.color = "black",
    table.border.bottom.color = "black"
  ) %>%
  tab_spanner(
    label = "Data Availability",
    columns = vars(race_available_percent, income_available_percent, tract_available_percent)
  ) %>%
  tab_spanner(
    label = "Summary Statistics (Median, Min, Max)",
    columns = vars(race_stats, income_stats, tract_stats)
  )


```

Medians

```{r}
olddata = haven::read_sas("P:/CH-Ranking/Data/2024/6 Measure Datasets/Additional Measures/v153.sas7bdat")



# Filter the data for different variables where each is not NA
filtered_race <- hist_data %>%
  filter(!is.na(race_cis_sd_bgv_wt))

filtered_income <- hist_data %>%
  filter(!is.na(income_cis_sd_bgv_wt))

filtered_tract <- hist_data %>%
  filter(!is.na(tract_cis_sd_bgv_wt))

# Merge olddata with filtered hist_data by statecode and countycode
merged_race <- olddata %>%
  inner_join(filtered_race, by = c("statecode", "countycode"))

merged_income <- olddata %>%
  inner_join(filtered_income, by = c("statecode", "countycode"))

merged_tract <- olddata %>%
  inner_join(filtered_tract, by = c("statecode", "countycode"))

# Calculate the median of v153_rawvalue for each filtered dataset
median_race <- median(merged_race$v153_rawvalue, na.rm = TRUE)
median_income <- median(merged_income$v153_rawvalue, na.rm = TRUE)
median_tract <- median(merged_tract$v153_rawvalue, na.rm = TRUE)

# Create a table using gt to display the median values
median_table <- tibble(
  Variable = c("v153_rawvalue with race data", "v153_rawvalue with income data", "v153_rawvalue with tract data"),
  Median_Value = c(median_race, median_income, median_tract)
) %>%
  gt() %>%
  tab_header(
    title = "Median Values of v153_rawvalue with available data"
  ) %>% 
  cols_label(
    Variable = "Variable",
    Median_Value = "Median Value"
  ) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black", weight = px(2)),
    locations = cells_body()
  ) %>%
  tab_options(
    table.width = pct(70),
    column_labels.font.size = 14,
    table.border.top.color = "black",
    table.border.bottom.color = "black"
  )

# Display the table
median_table

```
