---
title: "v153 Homeownership - disparity calculations"
author: "GL (edited on 3/19 by how)"
date: "2024-05-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidycensus)
```

# about

v153 Homeownership

numerator = [B25003_002E] Estimate; Total: - Owner occupied

denominator = [B25003_001E] Estimate; Total:

153	Homeownership	v153	NA	B25003	B25003_001	N	Estimate!!Total:	TENURE
153	Homeownership	v153	NA	B25003	B25003_002	N	Estimate!!Total:!!Owner occupied	TENURE


disparity calculations:

-   census tract 

-   race groups

-   income levels


# BGV

$$BGV=\sum_{i=1}^{n}p_i(y_i-\mu)^2$$

# SD

## Population standard deviation
If the data is being considered a population on its own, we divide by the number of data points, $N$

$$\sigma=\sqrt\frac{\sum_{i=1}^{N} (x_i - \mu)^2}{N}$$

## Sample standard deviation

If the data is a sample from a larger population, we divide by one fewer than the number of data points in the sample, $n-1$

$$s_x=\sqrt\frac{\sum_{i=1}^{n} (x_i - \mu)^2}{n-1}$$

# Theil

$$T=\frac{1}{n}\sum_{i=i}^{n}\frac{y_i}{\mu}log(\frac{y_i}{\mu})$$

## function: theil_wtd

```{r}
theil_wtd_gl <- function(x, weights=NULL)
{
  if (is.null(weights)){
    weights <- rep(1, length(x))}

  missing <- !(is.na(x) | is.nan(x) | is.na(weights))
  x <- x[missing]
  weights <- weights[missing]
  if (!all(weights>=0)) stop("At least one weight is negative", call.=FALSE)
  # if (all(weights == 0)) stop("All weights are zero", call.=FALSE)

  if (all(weights == 0)){return(NA_real_)}
  else {
    x_sel <- x[x>0]
    weights_sel <- weights[x>0]
    weights_sel <- weights_sel/sum(weights_sel)
    mean <- stats::weighted.mean(x_sel, weights_sel)
    x_sel <- x_sel/mean
    if(length(x_sel)==0){return(NA_real_)} # case for empty vector
    else{
      theil <- (sum(weights_sel*x_sel*log(x_sel)))
      return(theil)}
    
  }
  
}
```

```{r}
# test function
x <- c(0,1.5,2, NaN)
theil_wtd_gl(x)
```

# 1 tracts

## ACS variables 2022

### Detailed Tables (start with B or C)
```{r}

## Detailed Tables (start with B or C)
v22 <- load_variables(year = 2022, "acs5", cache = TRUE)

vars_B25003 <- v22 %>% 
  filter(str_detect(name, 'B25003'))

vars_B25003
```

### S: Subject tables (start with S)
```{r}
## S: Subject tables (start with S)
v22_s <- load_variables(year = 2022, dataset = "acs5/subject", cache = TRUE)

v22_s %>% 
  filter(str_detect(name, "S1901"))
```


### DP: Demographic Profile
```{r}
dp22 <- load_variables(2022, "acs5/profile", cache = TRUE)

dp22 %>% 
  filter(str_detect(name, "DP05"), str_detect(label, regex("Black", ignore_case=TRUE)))
```


## get tract data from Census

```{r}
# list of state abbreviations
sort(c(state.abb, "DC"))
```

### funciton

```{r}
# get tract level data for all states --------
get_acs_tract_data <- function(year = year, variables = variables, survey = "acs5") {
  # tract data
  print(paste0("Tract-level data: "))
  
  # sort(c(state.abb, "DC"))
  
  data_tract <- sort(c(state.abb, "DC")) %>% 
    map(~get_acs(state = .,
                 geography = "tract",
                 variables = variables,
                 year = year,
                 survey = survey,
                 output = "wide"))%>% 
    bind_rows() %>% 
    mutate(fipscode = str_sub(GEOID, 1L, 5L),
           statecode = str_sub(GEOID, 1L, 2L),
           countycode = str_sub(GEOID, 3L, 5L),
           .after = "GEOID")  
  
  return(data_tract)
}
```

### data

```{r}
acs_2023_v153_tract <- 
  get_acs_tract_data(year = 2023, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2022_v153_tract <- 
  get_acs_tract_data(year = 2022, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2021_v153_tract <- 
  get_acs_tract_data(year = 2021, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2020_v153_tract <- 
  get_acs_tract_data(year = 2020, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2019_v153_tract <- 
  get_acs_tract_data(year = 2019, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2018_v153_tract <- 
  get_acs_tract_data(year = 2018, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2017_v153_tract <- 
  get_acs_tract_data(year = 2017, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2016_v153_tract <- 
  get_acs_tract_data(year = 2016, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2015_v153_tract <- 
  get_acs_tract_data(year = 2015, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2014_v153_tract <- 
  get_acs_tract_data(year = 2014, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

acs_2013_v153_tract <- 
  get_acs_tract_data(year = 2013, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()

```

## efficient calcs of all years
```{r efficient calcs of all years with cis}


library(dplyr)
library(purrr)

# Create a named list of your datasets
acs_datasets <- list(
  `2013` = acs_2013_v153_tract,
  `2014` = acs_2014_v153_tract,
  `2015` = acs_2015_v153_tract,
  `2016` = acs_2016_v153_tract,
  `2017` = acs_2017_v153_tract,
  `2018` = acs_2018_v153_tract,
  `2019` = acs_2019_v153_tract,
  `2020` = acs_2020_v153_tract,
  `2021` = acs_2021_v153_tract,
  `2022` = acs_2022_v153_tract,
  `2023` = acs_2023_v153_tract
)

# Define the function to process each dataset
process_acs <- function(df) {
  df %>%
    rename(
      pop = B01001_001E,
      numerator = B25003_002E,
      denominator = B25003_001E,
      numerator_moe = B25003_002M,
      denominator_moe = B25003_001M
    ) %>%
    mutate(rawvalue = numerator / denominator) %>%
    mutate(
      se_rawvalue = sqrt((numerator_moe^2 / denominator^2) +
                         (numerator^2 * denominator_moe^2 / denominator^4)),
      ci_lower = rawvalue - 1.96 * se_rawvalue,
      ci_upper = rawvalue + 1.96 * se_rawvalue,
      rawvalue = if_else(ci_lower >= 0 & ci_upper <= 1 &
                         !is.na(ci_lower) & !is.na(ci_upper), rawvalue, NA_real_),
      pop_wt = if_else(is.na(rawvalue), 0, pop),
      pop_th = if_else(rawvalue > 0, pop, 0)
    ) %>%
    group_by(fipscode) %>%
    mutate(
      pop_i = pop_wt / sum(pop_wt),
      pop_th_i = pop_th / sum(pop_th),
      rawvalue_th = if_else(rawvalue == 0 | is.nan(rawvalue), NA_real_, rawvalue)
    ) %>%
    summarise(
      statecode = first(statecode), countycode = first(countycode),
      num_tract = n(),
      num_tract_counted = sum(!is.na(rawvalue)),
      pop_tot = sum(pop),
      pop_counted = sum(pop_wt),
      mean = mean(rawvalue, na.rm = TRUE),
      mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
      sd_s = sd(rawvalue, na.rm = TRUE),
      sd_p = sqrt((num_tract_counted - 1)/num_tract_counted) * sd(rawvalue, na.rm = TRUE),
      bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
      sd_bgv = sqrt(bgv),
      bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, na.rm = TRUE),
      sd_bgv_wt = sqrt(bgv_wt),
      num_tract_theil_counted = sum(rawvalue > 0, na.rm = TRUE),
      mean_th = mean(rawvalue_th, na.rm = TRUE),
      mean_th_wt = stats::weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
      theil = theil_wtd_gl(rawvalue),
      theil_wt = theil_wtd_gl(rawvalue, pop_wt),
      .groups = "drop"
    )
}

# Apply to all datasets and bind results
v153_disparity_tract_cis_all_years <- imap_dfr(
  acs_datasets,
  ~process_acs(.x) %>% mutate(year = .y)
)

```




# ----------

# 2 race groups: county level

-   8 groups

get v153 data by race: Tenure, 2022 ACS 5-Year Estimates, county level, tables: 
•	B25003B: Black
•	B25003C: AIAN
•	B25003D: ASIAN
•	B25003E: NHOPI
•	B25003F: SOR (Some Other Race)
•	B25003G: TOMR (Two or More Races)
•	B25003H: WHITE ALONE, Not Hispanic or Latino
•	B25003I: Hispanic
Numerators: [B25003B_002E] for Black through [B25003I_002E] for Hispanic
Denominators: [B25003B_001E] for Black through [B25003I_001E] for Hispanic


## function

```{r}
get_acs_data <- function(year = year, variables = variables, survey = "acs5") {
  # county data
  print("County-level data")
  cnt <- get_acs(
    geography = "county",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  )

  # state data
  print("State-level data")
  st <- get_acs(
    geography = "state",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  )

  # us data
  print("National data")
  us <- get_acs(
    geography = "us",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  ) %>% 
    mutate(GEOID = '0')

  # combine county, state, us data;
  data_all <- bind_rows(cnt, st, us) %>%
    arrange(GEOID) %>%
    # format GEOID;
    mutate(GEOID = str_pad(GEOID, 5, "right", "0")) %>%
    # add countycode, statecode;
    mutate(
      statecode = str_sub(GEOID, 1L, 2L),
      countycode = str_sub(GEOID, 3L, 5L),
      .after = "GEOID"
    ) %>%
    # remove PR
    filter(statecode != "72")
  
  return(data_all)
}
```

## data

### variables: v153 by race
```{r}
vars_B25003 %>% 
  slice(4:27) %>% 
  filter(!str_detect(name, "_003"))
```

```{r}
vars_v153_race <- vars_B25003 %>% 
               slice(4:27) %>% filter(!str_detect(name, "_003")) %>% 
               pull(name)

vars_v153_race
```

### variables: pop by race

```{r}
v22 %>% 
  filter(str_detect(name, "B01001"), str_detect(name, "_001")) %>% 
  slice(2:9)
```

```{r}
vars_pop_race <- v22 %>% 
  filter(str_detect(name, "B01001"), str_detect(name, "_001")) %>% 
  slice(2:9) %>% 
  pull(name)

vars_pop_race
```

### get data

```{r}
# county level
acs_2023_v153_race <- get_acs_data(year = 2023, variables = c(vars_v153_race,vars_pop_race) )
acs_2022_v153_race <- get_acs_data(year = 2022, variables = c(vars_v153_race,vars_pop_race) )
acs_2021_v153_race <- get_acs_data(year = 2021, variables = c(vars_v153_race,vars_pop_race) )
acs_2020_v153_race <- get_acs_data(year = 2020, variables = c(vars_v153_race,vars_pop_race) )
acs_2019_v153_race <- get_acs_data(year = 2019, variables = c(vars_v153_race,vars_pop_race) )
acs_2018_v153_race <- get_acs_data(year = 2018, variables = c(vars_v153_race,vars_pop_race) )

acs_2022_v153_race
```



## calculation

### numerator all years 

```{r}
# Create a named list of race datasets
acs_race_datasets <- list(
  `2018` = acs_2018_v153_race,
  `2019` = acs_2019_v153_race,
  `2020` = acs_2020_v153_race,
  `2021` = acs_2021_v153_race,
  `2022` = acs_2022_v153_race,
  `2023` = acs_2023_v153_race
)

# Function to process each year's race dataset
process_race_numerator <- function(df) {
  df_numerator <- df %>%
    select(GEOID, statecode, countycode, starts_with("B25003")) %>%
    filter(countycode != "000") %>%
    select(GEOID, statecode, countycode, ends_with("_002E")) %>%
    pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "numerator") %>%
    mutate(race = str_remove_all(race, "B25003|_002E"))
  
  df_moe <- df %>%
    select(GEOID, statecode, countycode, starts_with("B25003")) %>%
    filter(countycode != "000") %>%
    pivot_longer(cols = ends_with("_002M"), names_to = "race", values_to = "numerator_moe") %>%
    mutate(race = str_remove_all(race, "B25003|_002M"))
  
  df_numerator %>%
    left_join(df_moe, by = c("GEOID", "statecode", "countycode", "race"))
}

# Apply the function across all datasets and add year column
race_numerators_all_years <- imap_dfr(
  acs_race_datasets,
  ~process_race_numerator(.x) %>% mutate(year = .y)
)
```


### denominators all years 

```{r}


# Function to extract denominators and MOEs
process_race_denominator <- function(df) {
  df_denominator <- df %>%
    select(GEOID, statecode, countycode, starts_with("B25003")) %>%
    filter(countycode != "000") %>%
    select(GEOID, statecode, countycode, ends_with("_001E")) %>%
    pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "denominator") %>%
    mutate(race = str_remove_all(race, "B25003|_001E"))
  
  df_moe <- df %>%
    select(GEOID, statecode, countycode, starts_with("B25003")) %>%
    filter(countycode != "000") %>%
    pivot_longer(cols = ends_with("_001M"), names_to = "race", values_to = "denominator_moe") %>%
    mutate(race = str_remove_all(race, "B25003|_001M"))
  
  df_denominator %>%
    left_join(df_moe, by = c("GEOID", "statecode", "countycode", "race"))
}

# Apply function to all datasets, add year
race_denominators_all_years <- imap_dfr(
  acs_race_datasets,
  ~process_race_denominator(.x) %>% mutate(year = .y)
)

```


### pop all years 

```{r}
# Function to extract population by race
process_race_population <- function(df) {
  df %>%
    select(GEOID, statecode, countycode, starts_with("B01001")) %>%
    filter(countycode != "000") %>%
    select(GEOID, statecode, countycode, ends_with("_001E")) %>%
    pivot_longer(cols = starts_with("B01001"), names_to = "race", values_to = "pop") %>%
    mutate(race = str_remove_all(race, "B01001|_001E"))
}

# Apply function to all datasets, add year
race_population_all_years <- imap_dfr(
  acs_race_datasets,
  ~process_race_population(.x) %>% mutate(year = .y)
)

```


### disparity all years 


```{r including cis}
# Join all data together
v153_disparity_race_cis_all_years <- list(
  race_population_all_years,
  race_numerators_all_years,
  race_denominators_all_years
) %>%
  purrr::reduce(left_join, by = c("year", "GEOID", "statecode", "countycode", "race")) %>%
  group_by(year, GEOID) %>%
  mutate(
    rawvalue = numerator / denominator,
    ci_lower = (numerator - numerator_moe) / (denominator + denominator_moe), 
    ci_upper = (numerator + numerator_moe) / (denominator - denominator_moe),
    rawvalue = if_else(ci_lower < 0 | ci_upper > 1, NA_real_, rawvalue),
    pop_wt = if_else(is.na(rawvalue), 0, pop),
    pop_th = if_else(rawvalue > 0, pop, 0),
    pop_i = pop_wt / sum(pop_wt), .after = pop,
    rawvalue_th = if_else(rawvalue == 0 | is.nan(rawvalue), NA_real_, rawvalue)
  ) %>%
  summarise(
    statecode = first(statecode),
    countycode = first(countycode),
    num_grp = n(),
    num_grp_counted = sum(!is.na(rawvalue)),
    pop_tot = sum(pop),
    pop_counted = sum(pop_wt),
    mean = mean(rawvalue, na.rm = TRUE),
    mean_wt = weighted.mean(rawvalue, pop_i, na.rm = TRUE),
    sd_s = sd(rawvalue, na.rm = TRUE),
    sd_p = sqrt((num_grp_counted - 1) / num_grp_counted) * sd(rawvalue, na.rm = TRUE),
    bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
    sd_bgv = sqrt(bgv),
    bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, na.rm = TRUE),
    sd_bgv_wt = sqrt(bgv_wt),
    num_tract_theil_counted = sum(rawvalue > 0, na.rm = TRUE),
    mean_th = mean(rawvalue_th, na.rm = TRUE),
    mean_th_wt = weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
    theil = theil_wtd_gl(rawvalue),
    theil_wt = theil_wtd_gl(rawvalue, pop_wt),
    .groups = "drop"
  )
```






# ----------

# 3 income levels: county

## household

[Household](https://www.census.gov/programs-surveys/cps/technical-documentation/subject-definitions.html#household)

-   household

A household consists of all the people who occupy a housing unit. A house, an apartment or other group of rooms, or a single room, is regarded as a housing unit when it is occupied or intended for occupancy as separate living quarters; that is, when the occupants do not live with any other persons in the structure and there is direct access from the outside or through a common hall.

A household includes the related family members and all the unrelated people, if any, such as lodgers, foster children, wards, or employees who share the housing unit. A person living alone in a housing unit, or a group of unrelated people sharing a housing unit such as partners or roomers, is also counted as a household. The count of households excludes group quarters. There are two major categories of households, "family" and "nonfamily". (See definitions of Family household and Nonfamily household).

-   Family household 

A family household is a household maintained by a householder who is in a family (as defined above), and includes any unrelated people (unrelated subfamily members and/or secondary individuals) who may be residing there. The number of family households is equal to the number of families. The count of family household members differs from the count of family members, however, in that the family household members include all people living in the household, whereas family members include only the householder and his/her relatives. See the definition of family.

-   Nonfamily household

Household, nonfamily
A nonfamily household consists of a householder living alone (a one-person household) or where the householder shares the home exclusively with people to whom he/she is not related.

## data

### varialbes: B25118


```{r}
income_owner <- v22 %>% 
  filter(str_detect(name, "B25118"), str_detect(label, "Owner"), str_detect(label, "\\$")) 

income_owner
```

```{r}
income_renter <- v22 %>% 
  filter(str_detect(name, "B25118"), str_detect(label, "Renter"), str_detect(label, "\\$")) 

income_renter
```

### get B25118

#### owner

```{r}
acs_2023_B25118_owner <- get_acs_data(year = 2023, 
                                      variables = c(income_owner$name))
acs_2022_B25118_owner <- get_acs_data(year = 2022, 
                                      variables = c(income_owner$name))
acs_2021_B25118_owner <- get_acs_data(year = 2021, 
                                      variables = c(income_owner$name))
acs_2020_B25118_owner <- get_acs_data(year = 2020, 
                                      variables = c(income_owner$name))
acs_2019_B25118_owner <- get_acs_data(year = 2019, 
                                      variables = c(income_owner$name))
acs_2018_B25118_owner <- get_acs_data(year = 2018, 
                                      variables = c(income_owner$name))


```


```{r numerator with moe all years}

process_owner_data <- function(df, year, income_owner_df) {
  
  # Numerator
  long_nomoe <- df %>%
    select(fipscode = 1, 2:3, starts_with("B25118")) %>%
    filter(countycode != "000") %>%
    select(1:3, ends_with("E")) %>%
    pivot_longer(cols = starts_with("B25118"), names_to = "name", values_to = "numerator") %>%
    mutate(name = str_remove(name, "E")) %>%
    left_join(
      income_owner_df %>%
        select(1:2) %>%
        mutate(label = str_remove(label, "Estimate!!Total:!!Owner occupied:!!")),
      by = "name"
    )
  
  # MOE
  long_moe <- df %>%
    select(fipscode = 1, 2:3, starts_with("B25118")) %>%
    filter(countycode != "000") %>%
    select(1:3, ends_with("M")) %>%
    pivot_longer(cols = starts_with("B25118"), names_to = "name", values_to = "numerator_moe") %>%
    mutate(name = str_remove(name, "M")) %>%
    left_join(
      income_owner_df %>%
        select(1:2) %>%
        mutate(label = str_remove(label, "Estimate!!Total:!!Owner occupied:!!")),
      by = "name"
    )
  
  # Join numerator and MOE
  long_joined <- long_nomoe %>%
    left_join(long_moe, by = c("fipscode", "statecode", "countycode", "name", "label")) %>%
    mutate(year = year) %>%
    relocate(year)
  
  return(long_joined)
}


owner_datasets <- list(
  "2018" = acs_2018_B25118_owner,
  "2019" = acs_2019_B25118_owner,
  "2020" = acs_2020_B25118_owner,
  "2021" = acs_2021_B25118_owner,
  "2022" = acs_2022_B25118_owner,
  "2023" = acs_2023_B25118_owner
)
owner_all_years_long <- imap_dfr(owner_datasets, ~ process_owner_data(.x, .y, income_owner))



```



#### renter


```{r}
acs_2023_B25118_rent <- get_acs_data(year = 2023, 
                                     variables = c(income_renter$name))
acs_2022_B25118_rent <- get_acs_data(year = 2022, 
                                     variables = c(income_renter$name))
acs_2021_B25118_rent <- get_acs_data(year = 2021, 
                                     variables = c(income_renter$name))
acs_2020_B25118_rent <- get_acs_data(year = 2020, 
                                     variables = c(income_renter$name))
acs_2019_B25118_rent <- get_acs_data(year = 2019, 
                                     variables = c(income_renter$name))
acs_2018_B25118_rent <- get_acs_data(year = 2018, 
                                     variables = c(income_renter$name))

```

```{r denominator with moe all years }

process_renter_data <- function(df, year, income_renter_df) {
  
  # Denominator (estimates)
  long_nomoe <- df %>%
    select(fipscode = 1, 2:3, starts_with("B25118")) %>%
    filter(countycode != "000") %>%
    select(1:3, ends_with("E")) %>%
    pivot_longer(cols = starts_with("B25118"), names_to = "name", values_to = "denominator") %>%
    mutate(name = str_remove(name, "E")) %>%
    left_join(
      income_renter_df %>%
        select(1:2) %>%
        mutate(label = str_remove(label, "Estimate!!Total:!!Renter occupied:!!")),
      by = "name"
    )
  
  # MOE
  long_moe <- df %>%
    select(fipscode = 1, 2:3, starts_with("B25118")) %>%
    filter(countycode != "000") %>%
    select(1:3, ends_with("M")) %>%
    pivot_longer(cols = starts_with("B25118"), names_to = "name", values_to = "denominator_moe") %>%
    mutate(name = str_remove(name, "M")) %>%
    left_join(
      income_renter_df %>%
        select(1:2) %>%
        mutate(label = str_remove(label, "Estimate!!Total:!!Renter occupied:!!")),
      by = "name"
    )
  
  # Join denominator and MOE
  long_joined <- long_nomoe %>%
    left_join(long_moe, by = c("fipscode", "statecode", "countycode", "name", "label")) %>%
    mutate(year = year) %>%
    relocate(year)
  
  return(long_joined)
}

renter_datasets <- list(
  "2018" = acs_2018_B25118_rent,
  "2019" = acs_2019_B25118_rent,
  "2020" = acs_2020_B25118_rent,
  "2021" = acs_2021_B25118_rent,
  "2022" = acs_2022_B25118_rent,
  "2023" = acs_2023_B25118_rent
)

renter_all_years_long <- imap_dfr(renter_datasets, ~ process_renter_data(.x, .y, income_renter))



```


### numerator, denominator
```{r num denom with cis for all years }

compute_income_ci_by_group <- function(owner_df, renter_df, year) {
  owner_df %>%
    select(-name) %>%
    left_join(
      renter_df %>% select(-name),
      by = c("fipscode", "statecode", "countycode", "label")
    ) %>%
    select(1:4, label, numerator, denominator, numerator_moe, denominator_moe) %>%
    mutate(
      denominator = numerator + denominator,
      rawvalue = numerator / denominator,
      ci_lower = (numerator - numerator_moe) / (denominator + denominator_moe),
      ci_upper = (numerator + numerator_moe) / (denominator - denominator_moe),
      grp = case_when(
        label == 'Less than $5,000'         ~ 1,
        label == '$5,000 to $9,999'         ~ 1,
        label == '$10,000 to $14,999'       ~ 2,
        label == '$15,000 to $19,999'       ~ 3,
        label == '$20,000 to $24,999'       ~ 3,
        label == '$25,000 to $34,999'       ~ 4,
        label == '$35,000 to $49,999'       ~ 5,
        label == '$50,000 to $74,999'       ~ 6,
        label == '$75,000 to $99,999'       ~ 7,
        label == '$100,000 to $149,999'     ~ 8,
        label == '$150,000 or more'         ~ 9,
        TRUE ~ NA_integer_
      )
    ) %>%
    group_by(fipscode, statecode, countycode, grp) %>%
    summarise(
      numerator = sum(numerator, na.rm = TRUE),
      denominator = sum(denominator, na.rm = TRUE),
      ci_lower = sum(numerator - numerator_moe, na.rm = TRUE) / sum(denominator + denominator_moe, na.rm = TRUE),
       ci_upper = sum(numerator + numerator_moe, na.rm = TRUE) / sum(denominator - denominator_moe, na.rm = TRUE),
      rawvalue = if_else(
        ci_lower < 0 | ci_upper > 1,
        NA_real_,
        sum(numerator, na.rm = TRUE) / sum(denominator, na.rm = TRUE)
      ),
      .groups = "drop"
    ) %>%
    mutate(year = year) %>%
    relocate(year)
}

owner_list <- split(owner_all_years_long, as.character(owner_all_years_long$year))
renter_list <- split(renter_all_years_long, as.character(renter_all_years_long$year))

library(purrr)

income_ci_all_years <- imap_dfr(owner_list, function(owner_df, year) {
 # owner_df <- owner_list[[as.character(year)]]  # or owner_df <- owner_df[[1]]
  renter_df <- renter_list[[as.character(year)]]
  compute_income_ci_by_group(owner_df, renter_df, year)
})



```



### variables: S1901

S1901_C01_001	Estimate!!Households!!Total
S1901_C02_001	Estimate!!Families!!Total
S1901_C04_001	Estimate!!Nonfamily households!!Total
```{r}
test <- get_acs_data(year = yr, 
                     variables = c('S1901_C01_001', 'S1901_C02_001', 'S1901_C04_001'))

test %>% 
  select(-ends_with("M"))
```

### hoursefold: S1901
```{r}
S1901_household <- v22_s %>% 
  filter(str_detect(name, "S1901")) %>% 
  slice(1:11)

S1901_household %>% select(1:2)
```


```{r}
income_household_2023 <- get_acs_data(year = 2023, variables = c(S1901_household$name))
income_household_2022 <- get_acs_data(year = 2022, variables = c(S1901_household$name))
income_household_2021 <- get_acs_data(year = 2021, variables = c(S1901_household$name))
income_household_2020 <- get_acs_data(year = 2020, variables = c(S1901_household$name))
income_household_2019 <- get_acs_data(year = 2019, variables = c(S1901_household$name))
income_household_2018 <- get_acs_data(year = 2018, variables = c(S1901_household$name))


```

```{r hhi for all income groups }

process_income_household <- function(income_household, S1901_household) {
  income_household %>% 
    select(fipscode = 1, 2:3, starts_with('S1901')) %>% 
    filter(countycode != "000") %>% 
    select(1:3, ends_with('E')) %>% 
    mutate(across(S1901_C01_002E:S1901_C01_011E, ~round(./100 * S1901_C01_001E))) %>% 
    pivot_longer(cols = starts_with('S1901'), names_to = "name", values_to = 'num_household') %>% 
    mutate(name = str_remove(name, "E"),
           year = as.character(year)) %>% 
    left_join(
      S1901_household %>% 
        select(1:2) %>% 
        mutate(label = str_remove(label, "Estimate!!Households!!Total!!")),
      by = "name"
    ) %>% 
    filter(label != "Estimate!!Households!!Total") %>% 
    select(-name) %>% 
    mutate(grp = case_when(
      label == 'Less than $10,000'      ~ 1,  
      label == '$10,000 to $14,999'     ~ 2,  
      label == '$15,000 to $24,999'     ~ 3,  
      label == '$25,000 to $34,999'     ~ 4,  
      label == '$35,000 to $49,999'     ~ 5,  
      label == '$50,000 to $74,999'     ~ 6,  
      label == '$75,000 to $99,999'     ~ 7,  
      label == '$100,000 to $149,999'   ~ 8,      
      label == '$150,000 to $199,999'   ~ 9,
      label == '$200,000 or more'       ~ 9,
      TRUE ~ NA_integer_
    ), .after = label) %>% 
    group_by(fipscode, statecode, countycode, grp) %>% 
    summarise(num_household = sum(num_household, na.rm = TRUE), .groups = "drop")
}


income_household_list <- list(
  "2018" = income_household_2018,
  "2019" = income_household_2019,
  "2020" = income_household_2020,
  "2021" = income_household_2021,
  "2022" = income_household_2022,
  "2023" = income_household_2023
)

processed_list <- imap(income_household_list, function(df, year) {
  processed_df <- process_income_household(df, S1901_household) %>%
    mutate(year = as.integer(year), .before = 1)
  return(processed_df)
})

income_household_all_years <- bind_rows(processed_list)

```




### disparity

```{r income with cis all years }
income_ci_all_years$year = as.integer(income_ci_all_years$year)

v153_disparity_income_cis_allyears <- dplyr::left_join(
  income_household_all_years,
  income_ci_all_years,
  by = c('fipscode', 'statecode', 'countycode', 'grp', 'year')
) %>% 
  mutate(num_hh_wt = if_else(is.na(rawvalue), 0, num_household), .after = 'num_household') %>% 
  mutate(num_hh_th = if_else(rawvalue>0 & !is.na(rawvalue), num_household, 0)) %>% # set num_household=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode, year) %>% 
  mutate(hh_i = num_hh_wt / sum(num_hh_wt), .after = num_hh_wt) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_hh = sum(num_household), 
            num_hh_counted = sum(num_hh_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, hh_i, na.rm = TRUE),
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_hh_counted - 1)/num_hh_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(hh_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(hh_i * (rawvalue - weighted.mean(rawvalue, w = hh_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_hh_theil_counted = sum(num_hh_th, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, num_hh_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, num_hh_th),
            )
```



```{r with cis }
#note the rawvals are calcd above in the same step where the cis are calcd 
# therefore, rawvals are already available in acs_2022_v153_income_num_den_cis 
v153_disparity_income_cis <- purrr::reduce(
  list(acs_2022_v153_income_num_hshd, acs_2022_v153_income_num_den_cis),
  dplyr::left_join,
  by = c('fipscode', 'statecode', 'countycode', 'grp')
) %>% 
  mutate(num_hh_wt = if_else(is.na(rawvalue), 0, num_household), .after = 'num_household') %>% 
  mutate(num_hh_th = if_else(rawvalue>0 & !is.na(rawvalue), num_household, 0)) %>% # set num_household=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode) %>% 
  mutate(hh_i = num_hh_wt / sum(num_hh_wt), .after = num_hh_wt) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_hh = sum(num_household), 
            num_hh_counted = sum(num_hh_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, hh_i, na.rm = TRUE),
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_hh_counted - 1)/num_hh_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(hh_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(hh_i * (rawvalue - weighted.mean(rawvalue, w = hh_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_hh_theil_counted = sum(num_hh_th, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, num_hh_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, num_hh_th),
            )
```

```{r get the NAs corrected }
# Set specific columns to NA if they are 0
v153_disparity_tract_cis_all_years <- v153_disparity_tract_cis_all_years %>%
  mutate(sd_bgv_wt = ifelse( sd_bgv_wt == 0, NA, sd_bgv_wt))  
v153_disparity_race_cis_all_years <- v153_disparity_race_cis_all_years %>%
  mutate(sd_bgv_wt = ifelse( sd_bgv_wt == 0, NA, sd_bgv_wt))  
v153_disparity_income_cis_all_years <- v153_disparity_income_cis_allyears %>%
  mutate(sd_bgv_wt = ifelse( sd_bgv_wt == 0, NA, sd_bgv_wt))  



```



# save outputs 
```{r}
#save(v153_disparity_income, file = "calcd_data_how/income.RData")
#save(v153_disparity_income_cis, file = "calcd_data_how/income_cis.RData")
#save(v153_disparity_race, file = "calcd_data_how/race.RData")
#save(v153_disparity_race_cis, file = "calcd_data_how/race_cis.RData")
#save(v153_disparity_tract, file = "calcd_data_how/tract.RData")
#save(v153_disparity_tract_cis, file = "calcd_data_how/tract_cis.RData")

save(v153_disparity_tract_cis_all_years, file = "calcd_data_how/tract_cis_all_years.RData")
save(v153_disparity_income_cis_all_years, file = "calcd_data_how/income_cis_all_years.RData")
save(v153_disparity_race_cis_all_years, file = "calcd_data_how/race_cis_all_years.RData")

```










