---
title: "v153 Homeownership - disparity calculations"
author: "GL (edited on 3/19 by how)"
date: "2024-05-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidycensus)
```

# about

v153 Homeownership

numerator = [B25003_002E] Estimate; Total: - Owner occupied

denominator = [B25003_001E] Estimate; Total:

153	Homeownership	v153	NA	B25003	B25003_001	N	Estimate!!Total:	TENURE
153	Homeownership	v153	NA	B25003	B25003_002	N	Estimate!!Total:!!Owner occupied	TENURE


disparity calculations:

-   census tract 

-   race groups

-   income levels


# BGV

$$BGV=\sum_{i=1}^{n}p_i(y_i-\mu)^2$$

# SD

## Population standard deviation
If the data is being considered a population on its own, we divide by the number of data points, $N$

$$\sigma=\sqrt\frac{\sum_{i=1}^{N} (x_i - \mu)^2}{N}$$

## Sample standard deviation

If the data is a sample from a larger population, we divide by one fewer than the number of data points in the sample, $n-1$

$$s_x=\sqrt\frac{\sum_{i=1}^{n} (x_i - \mu)^2}{n-1}$$

# Theil

$$T=\frac{1}{n}\sum_{i=i}^{n}\frac{y_i}{\mu}log(\frac{y_i}{\mu})$$

## function: theil_wtd

```{r}
theil_wtd_gl <- function(x, weights=NULL)
{
  if (is.null(weights)){
    weights <- rep(1, length(x))}

  missing <- !(is.na(x) | is.nan(x) | is.na(weights))
  x <- x[missing]
  weights <- weights[missing]
  if (!all(weights>=0)) stop("At least one weight is negative", call.=FALSE)
  # if (all(weights == 0)) stop("All weights are zero", call.=FALSE)

  if (all(weights == 0)){return(NA_real_)}
  else {
    x_sel <- x[x>0]
    weights_sel <- weights[x>0]
    weights_sel <- weights_sel/sum(weights_sel)
    mean <- stats::weighted.mean(x_sel, weights_sel)
    x_sel <- x_sel/mean
    if(length(x_sel)==0){return(NA_real_)} # case for empty vector
    else{
      theil <- (sum(weights_sel*x_sel*log(x_sel)))
      return(theil)}
    
  }
  
}
```

```{r}
# test function
x <- c(0,1.5,2, NaN)
theil_wtd_gl(x)
```

# 1 tracts

## ACS variables 2022

### Detailed Tables (start with B or C)
```{r}
yr <- 2022

## Detailed Tables (start with B or C)
v22 <- load_variables(year = yr, "acs5", cache = TRUE)

vars_B25003 <- v22 %>% 
  filter(str_detect(name, 'B25003'))

vars_B25003
```

### S: Subject tables (start with S)
```{r}
## S: Subject tables (start with S)
v22_s <- load_variables(year = yr, dataset = "acs5/subject", cache = TRUE)

v22_s %>% 
  filter(str_detect(name, "S1901"))
```


### DP: Demographic Profile
```{r}
dp22 <- load_variables(2022, "acs5/profile", cache = TRUE)

dp22 %>% 
  filter(str_detect(name, "DP05"), str_detect(label, regex("Black", ignore_case=TRUE)))
```


## get tract data from Census

```{r}
# list of state abbreviations
sort(c(state.abb, "DC"))
```

### funciton

```{r}
# get tract level data for all states --------
get_acs_tract_data <- function(year = year, variables = variables, survey = "acs5") {
  # tract data
  print(paste0("Tract-level data: "))
  
  # sort(c(state.abb, "DC"))
  
  data_tract <- sort(c(state.abb, "DC")) %>% 
    map(~get_acs(state = .,
                 geography = "tract",
                 variables = variables,
                 year = yr,
                 survey = survey,
                 output = "wide"))%>% 
    bind_rows() %>% 
    mutate(fipscode = str_sub(GEOID, 1L, 5L),
           statecode = str_sub(GEOID, 1L, 2L),
           countycode = str_sub(GEOID, 3L, 5L),
           .after = "GEOID")  
  
  return(data_tract)
}
```

### data

```{r}
acs_2022_v153_tract <- 
  get_acs_tract_data(year = yr, variables = c('B25003_001', 'B25003_002', 'B01001_001')) %>% 
  glimpse()
```




## calculation

```{r withcis }
v153_disparity_tract_cis <- acs_2022_v153_tract %>% 
  #select(statecode, countycode, fipscode, B01001_001E, B25003_002E, B25003_001E, B25003_002M, B25003_001M) %>% 
  rename(pop = B01001_001E, 
         numerator = B25003_002E, 
         denominator = B25003_001E, 
         numerator_moe = B25003_002M, 
         denominator_moe = B25003_001M) %>% 
  mutate(rawvalue = numerator / denominator) %>% 
   # Calculate the lower and upper bounds for 95% CI of rawvalue
  mutate(
    se_rawvalue = sqrt((numerator_moe^2 / denominator^2) + 
                       (numerator^2 * denominator_moe^2 / denominator^4)),
    ci_lower = rawvalue - 1.96 * se_rawvalue,
    ci_upper = rawvalue + 1.96 * se_rawvalue
  ) %>% 
  # edit rawvalue column based on the condition
  mutate(rawvalue = if_else(ci_lower >= 0 & ci_upper <= 1 & !is.na(ci_lower) & !is.na(ci_upper), 
                               rawvalue, NA_real_)) %>% 
  mutate(pop_wt = if_else(is.na(rawvalue), 0, pop)) %>% # set pop=0 if a tract has missing raw value, i.e, exclude a tract in weighting if it has missing raw value
  mutate(pop_th = if_else(rawvalue>0, pop, 0)) %>% # set pop=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode) %>% 
  # mutate(pop_i = pop / sum(pop), .after = pop) %>% 
  mutate(pop_i = pop_wt / sum(pop_wt), .after = pop) %>% 
  mutate(pop_th_i = pop_th / sum(pop_th)) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  # filter(fipscode == "15005") %>%
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_tract = n(), 
            num_tract_counted = sum(!is.na(rawvalue)),
            pop_tot = sum(pop), 
            pop_counted = sum(pop_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
            # mean_wt_2 = sum(pop*rawvalue, na.rm = TRUE)/sum(pop), # this includes tracts with raw value missing but pop is not 0, which causes trouble
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_tract_counted - 1)/num_tract_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_tract_theil_counted = sum(rawvalue>0, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, pop_wt),
            )

```




```{r no suppression}
v153_disparity_tract <- acs_2022_v153_tract %>% 
  #select(statecode, countycode, fipscode, B01001_001E, B25003_002E, B25003_001E, B25003_002M, B25003_001M) %>% 
  rename(pop = B01001_001E, 
         numerator = B25003_002E, 
         denominator = B25003_001E, 
         numerator_moe = B25003_002M, 
         denominator_moe = B25003_001M) %>% 
  mutate(rawvalue = numerator/denominator) %>% 
  mutate(pop_wt = if_else(is.na(rawvalue), 0, pop)) %>% # set pop=0 if a tract has missing raw value, i.e, exclude a tract in weighting if it has missing raw value
  mutate(pop_th = if_else(rawvalue>0, pop, 0)) %>% # set pop=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode) %>% 
  # mutate(pop_i = pop / sum(pop), .after = pop) %>% 
  mutate(pop_i = pop_wt / sum(pop_wt), .after = pop) %>% 
  mutate(pop_th_i = pop_th / sum(pop_th)) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  # filter(fipscode == "15005") %>%
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_tract = n(), 
            num_tract_counted = sum(!is.na(rawvalue)),
            pop_tot = sum(pop), 
            pop_counted = sum(pop_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
            # mean_wt_2 = sum(pop*rawvalue, na.rm = TRUE)/sum(pop), # this includes tracts with raw value missing but pop is not 0, which causes trouble
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_tract_counted - 1)/num_tract_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_tract_theil_counted = sum(rawvalue>0, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, pop_wt),
            )

v153_disparity_tract
```


# ----------

# 2 race groups: county level

-   8 groups

get v153 data by race: Tenure, 2022 ACS 5-Year Estimates, county level, tables: 
•	B25003B: Black
•	B25003C: AIAN
•	B25003D: ASIAN
•	B25003E: NHOPI
•	B25003F: SOR (Some Other Race)
•	B25003G: TOMR (Two or More Races)
•	B25003H: WHITE ALONE, Not Hispanic or Latino
•	B25003I: Hispanic
Numerators: [B25003B_002E] for Black through [B25003I_002E] for Hispanic
Denominators: [B25003B_001E] for Black through [B25003I_001E] for Hispanic


## function

```{r}
get_acs_data <- function(year = year, variables = variables, survey = "acs5") {
  # county data
  print("County-level data")
  cnt <- get_acs(
    geography = "county",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  )

  # state data
  print("State-level data")
  st <- get_acs(
    geography = "state",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  )

  # us data
  print("National data")
  us <- get_acs(
    geography = "us",
    variables = variables,
    year = year,
    survey = survey,
    output = "wide"
  ) %>% 
    mutate(GEOID = '0')

  # combine county, state, us data;
  data_all <- bind_rows(cnt, st, us) %>%
    arrange(GEOID) %>%
    # format GEOID;
    mutate(GEOID = str_pad(GEOID, 5, "right", "0")) %>%
    # add countycode, statecode;
    mutate(
      statecode = str_sub(GEOID, 1L, 2L),
      countycode = str_sub(GEOID, 3L, 5L),
      .after = "GEOID"
    ) %>%
    # remove PR
    filter(statecode != "72")
  
  return(data_all)
}
```

## data

### variables: v153 by race
```{r}
vars_B25003 %>% 
  slice(4:27) %>% 
  filter(!str_detect(name, "_003"))
```

```{r}
vars_v153_race <- vars_B25003 %>% 
               slice(4:27) %>% filter(!str_detect(name, "_003")) %>% 
               pull(name)

vars_v153_race
```

### variables: pop by race

```{r}
v22 %>% 
  filter(str_detect(name, "B01001"), str_detect(name, "_001")) %>% 
  slice(2:9)
```

```{r}
vars_pop_race <- v22 %>% 
  filter(str_detect(name, "B01001"), str_detect(name, "_001")) %>% 
  slice(2:9) %>% 
  pull(name)

vars_pop_race
```

### get data

```{r}
# county level
acs_2022_v153_race <- get_acs_data(year = yr, variables = c(vars_v153_race,vars_pop_race) )

acs_2022_v153_race
```


#### DP
DP05_0080	Estimate!!HISPANIC OR LATINO AND RACE!!Total population!!Not Hispanic or Latino!!Black or African American alone
DP05_0038	Estimate!!RACE!!Total population!!One race!!Black or African American
DP05_0067	Estimate!!Race alone or in combination with one or more other races!!Total population!!Black or African American


```{r}
pop_dp <- get_acs_data(year = yr, 
                       variables = c('DP05_0080','DP05_0038', 'DP05_0067') )

```

## calculation

### numerator

```{r}


acs_2022_v153_race_numerator <- acs_2022_v153_race %>% 
  dplyr::select(GEOID, statecode, countycode, starts_with('B25003')) %>% 
  filter(countycode != "000") %>% 
  dplyr::select(GEOID, statecode, countycode, ends_with('_002E')) %>% 
  pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "numerator")%>% 
  mutate(race = str_remove_all(race, "B25003|_002E"))

# Pivoting for the margin of error (ends with _001M)
acs_2022_v153_race_moe <- acs_2022_v153_race %>% 
  dplyr::select(GEOID, statecode, countycode, starts_with('B25003')) %>% 
  filter(countycode != "000") %>%
  pivot_longer(cols = ends_with('_002M'), names_to = "race", values_to = "numerator_moe") %>%
  mutate(race = str_remove_all(race, "B25003|_002M"))

# Joining the denominator and margin of error by GEOID, statecode, countycode, and race
acs_2022_v153_race_num <- acs_2022_v153_race_numerator %>%
  left_join(acs_2022_v153_race_moe, by = c("GEOID", "statecode", "countycode", "race"))

acs_2022_v153_race_num
```

### denominator
```{r}
acs_2022_v153_race_denominator <- acs_2022_v153_race %>% 
  dplyr::select(GEOID, statecode, countycode, starts_with('B25003')) %>% 
  filter(countycode != "000") %>% 
  dplyr::select(GEOID, statecode, countycode, ends_with('_001E')) %>% 
  pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "denominator") %>% 
  mutate(race = str_remove_all(race, "B25003|_001E"))


# Pivoting for the margin of error (ends with _001M)
acs_2022_v153_race_moe <- acs_2022_v153_race %>% 
  dplyr::select(GEOID, statecode, countycode, starts_with('B25003')) %>% 
  filter(countycode != "000") %>%
  pivot_longer(cols = ends_with('_001M'), names_to = "race", values_to = "denominator_moe") %>%
  mutate(race = str_remove_all(race, "B25003|_001M"))

# Joining the denominator and margin of error by GEOID, statecode, countycode, and race
acs_2022_v153_race_den <- acs_2022_v153_race_denominator %>%
  left_join(acs_2022_v153_race_moe, by = c("GEOID", "statecode", "countycode", "race"))


```

### pop

```{r}
acs_2022_v153_race_pop <- acs_2022_v153_race %>% 
  dplyr::select(GEOID, statecode, countycode, starts_with('B01001')) %>% 
  filter(countycode != "000") %>% 
  dplyr::select(GEOID, statecode, countycode, ends_with('_001E')) %>% 
  pivot_longer(cols = starts_with("B01001"), names_to = "race", values_to = "pop") %>% 
  mutate(race = str_remove_all(race, "B01001|_001E"))

acs_2022_v153_race_pop
```

### disparity

```{r}
v153_disparity_race <- purrr::reduce(
  list(acs_2022_v153_race_pop, acs_2022_v153_race_num, acs_2022_v153_race_den),
  dplyr::left_join,
  by = c('GEOID', 'statecode', 'countycode', 'race')
) %>% 
  mutate(rawvalue = numerator/denominator) %>% 
  mutate(pop_wt = if_else(is.na(rawvalue), 0, pop)) %>% # set pop=0 if a group has missing raw value, i.e, exclude a group in weighting if it has missing raw value
  mutate(pop_th = if_else(rawvalue>0, pop, 0)) %>% # set pop=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(GEOID) %>% 
  mutate(pop_i = pop_wt / sum(pop_wt), .after = pop) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_grp = n(), 
            num_grp_counted = sum(!is.na(rawvalue)),
            pop_tot = sum(pop), 
            pop_counted = sum(pop_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_grp_counted - 1)/num_grp_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_tract_theil_counted = sum(rawvalue>0, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, pop_wt),
            )

v153_disparity_race
```


```{r disparity race with cis}
v153_disparity_race_cis <- purrr::reduce(
  list(acs_2022_v153_race_pop, acs_2022_v153_race_num, acs_2022_v153_race_den),
  dplyr::left_join,
  by = c('GEOID', 'statecode', 'countycode', 'race')
) %>%
  group_by(GEOID) %>% 
  mutate(
    rawvalue = numerator / denominator,
    # Calculate the Lower and Upper Confidence Intervals (CIs)
    ci_lower = (numerator - numerator_moe) / (denominator + denominator_moe), 
    ci_upper = (numerator + numerator_moe) / (denominator - denominator_moe),
    
     # Set rawvalue to NA if ci_lower < 0 or ci_upper > 1
    rawvalue = if_else(ci_lower < 0 | ci_upper > 1, NA_real_, rawvalue),
    
    pop_wt = if_else(is.na(rawvalue), 0, pop), # set pop=0 if a group has missing raw value
    pop_th = if_else(rawvalue > 0, pop, 0), # set pop=0 if a tract has missing or 0 raw value
      pop_i = pop_wt / sum(pop_wt), .after = pop,
      rawvalue_th = if_else(rawvalue == 0 | is.nan(rawvalue), NA, rawvalue)
    ) %>% 
    summarise(
      statecode = first(statecode),
      countycode = first(countycode),
      num_grp = n(),
      num_grp_counted = sum(!is.na(rawvalue)),
      pop_tot = sum(pop),
      pop_counted = sum(pop_wt),
      mean = mean(rawvalue, na.rm = TRUE),
      mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
      sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
      sd_p = sqrt((num_grp_counted - 1) / num_grp_counted) * sd(rawvalue, na.rm = TRUE), # population SD
      bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
      sd_bgv = sqrt(bgv),
      bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, 
                   na.rm = TRUE),
      sd_bgv_wt = sqrt(bgv_wt),
      num_tract_theil_counted = sum(rawvalue > 0, na.rm = TRUE),
      mean_th = mean(rawvalue_th, na.rm = TRUE),
      mean_th_wt = stats::weighted.mean(rawvalue_th, pop_th, na.rm = TRUE),
      theil = theil_wtd_gl(rawvalue),
      theil_wt = theil_wtd_gl(rawvalue, pop_wt)
    )
  

      
      
```



# ----------

# 2b. tracts + race

## data

```{r}
vars_v153_race

vars_pop_race
```

```{r}
# tract level
acs_2022_v153_race_tract <- get_acs_tract_data(year = yr, variables = c(vars_v153_race,vars_pop_race))

acs_2022_v153_race_tract
```


## calculation
### numerator
```{r}
acs_2022_v153_race_tract_num <- acs_2022_v153_race_tract %>% 
  select(1:4, starts_with('B25003')) %>% 
  select(1:4, ends_with('_002E')) %>% 
  pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "numerator")%>% 
  mutate(race = str_remove_all(race, "B25003|_002E"))

acs_2022_v153_race_tract_num
```

### denominator
```{r}
acs_2022_v153_race_tract_den <- acs_2022_v153_race_tract %>% 
  select(1:4, starts_with('B25003')) %>% 
  select(1:4, ends_with('_001E')) %>% 
  pivot_longer(cols = starts_with("B25003"), names_to = "race", values_to = "denominator")%>% 
  mutate(race = str_remove_all(race, "B25003|_001E"))

acs_2022_v153_race_tract_den
```

### pop

```{r}
acs_2022_v153_race_tract_pop <- acs_2022_v153_race_tract %>% 
  select(1:4, starts_with('B01001')) %>% 
  select(1:4, ends_with('_001E')) %>% 
  pivot_longer(cols = starts_with("B01001"), names_to = "race", values_to = "pop") %>% 
  mutate(race = str_remove_all(race, "B01001|_001E"))

acs_2022_v153_race_tract_pop
```

### disparity

```{r}
v153_disparity_race_tract <- purrr::reduce(
  list(acs_2022_v153_race_tract_pop, acs_2022_v153_race_tract_num, acs_2022_v153_race_tract_den),
  dplyr::left_join,
  by = c('GEOID','fipscode', 'statecode', 'countycode', 'race')
) %>% 
  mutate(rawvalue = numerator/denominator) %>% 
  mutate(pop_wt = if_else(is.na(rawvalue), 0, pop)) %>% 
  group_by(fipscode) %>% 
  # mutate(pop_i = pop / sum(pop), .after = pop) %>% 
  mutate(pop_i = pop_wt / sum(pop_wt), .after = pop) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_grp = n(), 
            num_grp_counted = sum(!is.na(rawvalue)),
            pop_tot = sum(pop), 
            pop_counted = sum(pop_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, pop_i, na.rm = TRUE),
            # mean_wt_2 = sum(pop*rawvalue, na.rm = TRUE)/sum(pop), # this includes tracts with raw value missing but pop is not 0, which causes trouble
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            # sd =  sqrt((num_tracts - 1)/num_tracts) * sd(rawvalue, na.rm = TRUE), # population SD
            sd_p =  sqrt((num_grp_counted - 1)/num_grp_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(pop_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(pop_i * (rawvalue - weighted.mean(rawvalue, w = pop_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            )

v153_disparity_race_tract
```


# tract vs. race vs. tract+trace

```{r}
v153_disparity_tract %>% 
  select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
  rename_with(.fn = ~paste0(.x, "_Tract"), .cols = c("mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" ))

v153_disparity_race %>% 
  select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
  rename_with(.fn = ~paste0(.x, "_Race"), .cols = c("mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" ))

v153_disparity_race_tract %>% 
  select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
  rename_with(.fn = ~paste0(.x, "_RandT"), .cols = c("num_grp", "num_grp_counted", "mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" ))
```

```{r}
v153_disparity_TractRaceTandR <- purrr::reduce(
  list(
    v153_disparity_tract %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename(num_grp = num_tract, num_grp_counted = num_tract_counted) %>% 
      rename_with(.fn = ~paste0(.x, "_Tract"), .cols = c('num_grp', 'num_grp_counted',"mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" )),

    v153_disparity_race %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename_with(.fn = ~paste0(.x, "_Race"), .cols = c("num_grp", "num_grp_counted","mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" )),

    v153_disparity_race_tract %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename_with(.fn = ~paste0(.x, "_RandT"), .cols = c("num_grp", "num_grp_counted", "mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" ))
        
    ),
  dplyr::left_join,
  by = c('fipscode')
) %>% 
  left_join(fips, by = "fipscode") %>% 
  select(fipscode, statecode, countycode, state, county, everything())

v153_disparity_TractRaceTandR
```

```{r}
p <- v153_disparity_TractRaceTandR %>% 
  filter(state=="WI") %>% 
  select(-starts_with("num_grp"), -c(2:5)) %>% 
  pivot_longer(cols = c(mean_Tract:sd_bgv_wt_RandT), names_to = "name", values_to = "value") %>% 
  mutate(variable = str_remove_all(name, "_Tract|_Race|_RandT"),
         cat = case_when(
           str_detect(name, "_Tract$") ~ 1,
           str_detect(name, "_Race$") ~ 2,
           str_detect(name, "_RandT$") ~ 3,
         )
         ) %>% 
  filter(variable %in% c("mean")) %>% 
  ggplot(aes(fipscode, value, group = name, color = fct_reorder(name, cat) )) +
  geom_line() + geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.title=element_blank(),
        legend.position="bottom") +
  labs(title = "mean")

plotly::ggplotly(
  p
)
```


```{r}
c("_Tract", "_Race", "_RandT")

rep("mean", 3)

paste0("mean", c("_Tract", "_Race", "_RandT"))

eval(substitute('mean_Tract'), v153_disparity_TractRaceTandR)
```
```{r}
library(plotly)

st <- "WI"

stats <- "mean"

y_cols <- paste0(stats, c("_Tract", "_Race", "_RandT"))

plot_ly(v153_disparity_TractRaceTandR %>% 
          filter(state==st) %>% 
          select(-starts_with("num_grp")),
        x = ~fct_reorder(county, fipscode), y = ~ mean_Tract, name = "mean_Tract",
        type = 'scatter',mode = 'lines+markers'
  ) %>% 
  add_trace(
    y = ~mean_Race, name = "mean_Race"
  ) %>%
  add_trace(
    y = ~mean_RandT, name = "mean_RandT"
  ) %>%
  layout(title = paste0("State: ", st),
         xaxis = list(title = "County"),
         yaxis = list (title = stats))
```


```{r}
library(plotly)

st <- "WI"

stats <- "mean"

y_cols <- paste0(stats, c("_Tract", "_Race", "_RandT"))

plot_ly(v153_disparity_TractRaceTandR %>% 
          filter(state==st) %>% 
          select(-starts_with("num_grp"), -c(2:5)),
        x = ~fipscode, y = ~ select(.data, y_cols[1]) %>% pull() , name = y_cols[1],
        type = 'scatter',mode = 'lines+markers'
  ) %>% 
  add_trace(
    y = ~select(.data, y_cols[2]) %>% pull() , name = y_cols[2]
  ) %>%
  add_trace(
    y = ~select(.data, y_cols[3]) %>% pull() , name = y_cols[3]
  ) %>%
  layout(title = paste0("State: ", st),
         xaxis = list(title = "County"),
         yaxis = list (title = stats))
```

## funciton: plot a stats in a state

```{r}
plot_one_stats <- function(df, state_abbr, stats_sel ){

  y_cols <- paste0(stats_sel, c("_Tract", "_Race", "_RandT"))
  
  plot_ly(df %>% 
            filter(state==state_abbr) %>% 
            select(-starts_with("num_grp")),
          x = ~fct_reorder(county, fipscode), y = ~ select(.data, y_cols[1]) %>% pull() , name = y_cols[1],
          type = 'scatter',mode = 'lines+markers'
    ) %>% 
    add_trace(
      y = ~select(.data, y_cols[2]) %>% pull() , name = y_cols[2]
    ) %>%
    add_trace(
      y = ~select(.data, y_cols[3]) %>% pull() , name = y_cols[3]
    ) %>%
    layout(title = paste0("State: ", st),
           xaxis = list(title = "County"),
           yaxis = list (title = stats_sel))
}

plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "mean_wt" )
```


## mean
```{r}
# mean
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, mean_Tract, mean_Race, mean_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "mean") %>% 
  ggplot(aes(fipscode, mean, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "mean")
)

```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "mean" )
```


## mean_wt

-   weighted mean: align with county-level ranking data

```{r}
# mean_wt
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, mean_wt_Tract, mean_wt_Race, mean_wt_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "mean_wt") %>% 
  ggplot(aes(fipscode, mean_wt, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "weighted mean - WI")
)

```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "mean_wt" )
```

## bgv
```{r}
# bgv
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, bgv_Tract, bgv_Race, bgv_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "bgv") %>% 
  ggplot(aes(fipscode, bgv, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "BGV - WI")
)
```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "bgv" )
```

## sd_bgv
```{r}
# bgv
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, sd_bgv_Tract, sd_bgv_Race, sd_bgv_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "sd_bgv") %>% 
  ggplot(aes(fipscode, sd_bgv, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
)
```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "sd_bgv" )
```

## bgv_wt
```{r}
# bgv_wt
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, bgv_wt_Tract, bgv_wt_Race, bgv_wt_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "bgv_wt") %>% 
  ggplot(aes(fipscode, bgv_wt, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "weighted BGV - WI")
  
)
```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "bgv_wt" )
```

## sd_bgv_wt

```{r}
# bgv_wt
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>%
  select(1, sd_bgv_wt_Tract, sd_bgv_wt_Race, sd_bgv_wt_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "sd_bgv_wt") %>% 
  ggplot(aes(fipscode, sd_bgv_wt, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
)
```

```{r}
plot_one_stats(df=v153_disparity_TractRaceTandR, state_abbr="WI", stats_sel = "sd_bgv_wt" )
```

## mean_wt vs. sd_bgv_wt

```{r}
# mean_wt vs. sd_bgv_wt
plotly::ggplotly(
  purrr::reduce(
  list(
    v153_disparity_tract %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename_with(.fn = ~paste0(.x, "_Tract"), .cols = c("mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" )),

    v153_disparity_race %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename_with(.fn = ~paste0(.x, "_Race"), .cols = c("mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" )),

    v153_disparity_race_tract %>% 
      select(-c(statecode, countycode, sd_s, sd_p), -starts_with("pop")) %>% 
      rename_with(.fn = ~paste0(.x, "_RandT"), .cols = c("num_grp", "num_grp_counted", "mean", "mean_wt", "bgv" , "sd_bgv", "bgv_wt", "sd_bgv_wt" ))
        
    ),
  dplyr::left_join,
  by = c('fipscode')
) %>% 
  filter(str_detect(fipscode, "^55")) %>%
  select(1, contains("_wt_")) %>% 
  select(-c(bgv_wt_Tract, bgv_wt_Race, bgv_wt_RandT)) %>% 
  pivot_longer(cols = c(mean_wt_Tract:sd_bgv_wt_RandT), names_to = "variable", values_to = "value") %>% 
  ggplot(aes(fipscode, value, group = variable, color = variable)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
)
```

# ----------

# 3 income levels: county

## household

[Household](https://www.census.gov/programs-surveys/cps/technical-documentation/subject-definitions.html#household)

-   household

A household consists of all the people who occupy a housing unit. A house, an apartment or other group of rooms, or a single room, is regarded as a housing unit when it is occupied or intended for occupancy as separate living quarters; that is, when the occupants do not live with any other persons in the structure and there is direct access from the outside or through a common hall.

A household includes the related family members and all the unrelated people, if any, such as lodgers, foster children, wards, or employees who share the housing unit. A person living alone in a housing unit, or a group of unrelated people sharing a housing unit such as partners or roomers, is also counted as a household. The count of households excludes group quarters. There are two major categories of households, "family" and "nonfamily". (See definitions of Family household and Nonfamily household).

-   Family household 

A family household is a household maintained by a householder who is in a family (as defined above), and includes any unrelated people (unrelated subfamily members and/or secondary individuals) who may be residing there. The number of family households is equal to the number of families. The count of family household members differs from the count of family members, however, in that the family household members include all people living in the household, whereas family members include only the householder and his/her relatives. See the definition of family.

-   Nonfamily household

Household, nonfamily
A nonfamily household consists of a householder living alone (a one-person household) or where the householder shares the home exclusively with people to whom he/she is not related.

## data

### varialbes: B25118


```{r}
income_owner <- v22 %>% 
  filter(str_detect(name, "B25118"), str_detect(label, "Owner"), str_detect(label, "\\$")) 

income_owner
```

```{r}
income_renter <- v22 %>% 
  filter(str_detect(name, "B25118"), str_detect(label, "Renter"), str_detect(label, "\\$")) 

income_renter
```

### get B25118

#### owner

```{r}
acs_2022_B25118_owner <- get_acs_data(year = yr, 
                                      variables = c(income_owner$name))

acs_2022_B25118_owner
```

```{r numerator with moe }

acs_2022_B25118_owner_long_nomoe <- acs_2022_B25118_owner %>% 
  select(fipscode = 1, 2:3, starts_with('B25118')) %>% 
  filter(countycode != "000") %>% 
  select(1:3, ends_with('E')) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'numerator') %>% 
  mutate(name = str_remove(name, "E")) %>% 
  left_join(income_owner %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Owner occupied:!!")),
            by = "name"
  )




acs_2022_B25118_owner_long_moe <- acs_2022_B25118_owner %>% 
  select(fipscode = 1, 2:3, starts_with('B25118')) %>% 
  filter(countycode != "000") %>% 
  select(1:3, ends_with('M')) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'numerator_moe') %>% 
  mutate(name = str_remove(name, "M")) %>% 
  left_join(income_owner %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Owner occupied:!!")),
            by = "name"
  )


# Joining the denominator and margin of error by fipscode, statecode, countycode, and variable
acs_2022_B25118_owner_long <- acs_2022_B25118_owner_long_nomoe %>%
  left_join(acs_2022_B25118_owner_long_moe, by = c("fipscode", "statecode", "countycode", "name", "label"))

```

#### renter


```{r}
acs_2022_B25118_rent <- get_acs_data(year = yr, 
                                     variables = c(income_renter$name))

acs_2022_B25118_rent
```

```{r denominator with moe}
acs_2022_B25118_rent_long_nomoe <- acs_2022_B25118_rent %>% 
  select(fipscode = 1, 2:3, starts_with('B25118')) %>% 
  filter(countycode != "000") %>% 
  select(1:3, ends_with('E')) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'denominator') %>% 
  mutate(name = str_remove(name, "E")) %>% 
  left_join(income_renter %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Renter occupied:!!")),
            by = "name"
  )


acs_2022_B25118_rent_long_moe <- acs_2022_B25118_rent %>% 
  select(fipscode = 1, 2:3, starts_with('B25118')) %>% 
  filter(countycode != "000") %>% 
  select(1:3, ends_with('M')) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'denominator_moe') %>% 
  mutate(name = str_remove(name, "M")) %>% 
  left_join(income_renter %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Renter occupied:!!")),
            by = "name"
  )

# Joining the denominator and margin of error by fipscode, statecode, countycode, and variable
acs_2022_B25118_rent_long <- acs_2022_B25118_rent_long_nomoe %>%
  left_join(acs_2022_B25118_rent_long_moe, by = c("fipscode", "statecode", "countycode", "name", "label"))


```

### numerator, denominator

```{r no cis}
acs_2022_v153_income_num_den <- acs_2022_B25118_owner_long %>% 
  select(-name) %>% 
  left_join(
    acs_2022_B25118_rent_long %>% 
      select(-name),
    by = c("fipscode", "statecode", "countycode", "label")
    ) %>% 
  select(1:3, label, numerator, denominator) %>% 
  # filter(is.na(numerator) | is.na(denominator)) %>% 
  mutate(denominator = numerator + denominator) %>% 
  mutate(grp = case_when(
    label =='Less than $5,000'		~ 1,	
    label =='$5,000 to $9,999'		~ 1,	
    label =='$10,000 to $14,999'	~ 2,	
    label =='$15,000 to $19,999'	~ 3,	
    label =='$20,000 to $24,999'	~ 3,	
    label =='$25,000 to $34,999'	~ 4,	
    label =='$35,000 to $49,999'	~ 5,	
    label =='$50,000 to $74,999'	~ 6,	
    label =='$75,000 to $99,999'	~ 7,	
    label =='$100,000 to $149,999'	~ 8,		
    label =='$150,000 or more'      ~ 9,
    TRUE ~ NA
  ), .after = label
  ) %>% 
  group_by(fipscode,statecode,countycode, grp) %>% 
  summarise(numerator = sum(numerator), 
            denominator = sum(denominator), .groups = "drop") 

acs_2022_v153_income_num_den
```

```{r with cis }

acs_2022_v153_income_num_den_cis <- acs_2022_B25118_owner_long %>% 
  select(-name) %>% 
  left_join(
    acs_2022_B25118_rent_long %>% 
      select(-name),
    by = c("fipscode", "statecode", "countycode", "label")
  ) %>% 
  select(1:3, label, numerator, denominator, numerator_moe, denominator_moe) %>%  # Include the margin of error columns
  mutate(
    denominator = numerator + denominator,
    rawvalue = numerator / denominator,  # Calculate rawvalue
    # Calculate the 95% CI for the rawvalue
    ci_lower = (numerator - numerator_moe) / (denominator + denominator_moe),  # Lower CI
    ci_upper = (numerator + numerator_moe) / (denominator - denominator_moe)   # Upper CI
  ) %>% 
   # Set rawvalue to NA if the lower CI is less than 0 or the upper CI is greater than 1
  #mutate(rawvalue = if_else(ci_lower < 0 | ci_upper > 1, NA_real_, rawvalue)
  #) %>% 
  mutate(grp = case_when(
    label == 'Less than $5,000'		~ 1,	
    label == '$5,000 to $9,999'		~ 1,	
    label == '$10,000 to $14,999'	~ 2,	
    label == '$15,000 to $19,999'	~ 3,	
    label == '$20,000 to $24,999'	~ 3,	
    label == '$25,000 to $34,999'	~ 4,	
    label == '$35,000 to $49,999'	~ 5,	
    label == '$50,000 to $74,999'	~ 6,	
    label == '$75,000 to $99,999'	~ 7,	
    label == '$100,000 to $149,999'	~ 8,		
    label == '$150,000 or more'      ~ 9,
    TRUE ~ NA
  ), .after = label) %>% 
  group_by(fipscode, statecode, countycode, grp) %>% 
  summarise(
    numerator = sum(numerator, na.rm = TRUE), 
    denominator = sum(denominator, na.rm = TRUE), 
    #rawvalue = sum(numerator, na.rm = TRUE) / sum(denominator, na.rm = TRUE),  # Summing up numerator and denominator for the group
    ci_lower = sum(numerator - numerator_moe, na.rm = TRUE) / sum(denominator + denominator_moe, na.rm = TRUE),  # Lower CI for the group
    ci_upper = sum(numerator + numerator_moe, na.rm = TRUE) / sum(denominator - denominator_moe, na.rm = TRUE), 
     rawvalue = if_else(ci_lower < 0 | ci_upper > 1, NA_real_, sum(numerator, na.rm = TRUE) / sum(denominator, na.rm = TRUE)),# Upper CI for the group
    .groups = "drop"
  )



```



### variables: S1901

S1901_C01_001	Estimate!!Households!!Total
S1901_C02_001	Estimate!!Families!!Total
S1901_C04_001	Estimate!!Nonfamily households!!Total
```{r}
test <- get_acs_data(year = yr, 
                     variables = c('S1901_C01_001', 'S1901_C02_001', 'S1901_C04_001'))

test %>% 
  select(-ends_with("M"))
```

### hoursefold: S1901
```{r}
S1901_household <- v22_s %>% 
  filter(str_detect(name, "S1901")) %>% 
  slice(1:11)

S1901_household %>% select(1:2)
```


```{r}
income_household <- get_acs_data(year = yr, variables = c(S1901_household$name))
```




#### method: GL
```{r}
#we don't need the moe here since these vals are just used for weights 

acs_2022_v153_income_num_hshd <- income_household %>% 
  select(fipscode = 1, 2:3, starts_with('S1901')) %>% 
  filter(countycode != "000") %>% 
  select(1:3, ends_with('E')) %>% 
  mutate(across(S1901_C01_002E:S1901_C01_011E, ~round(./100*S1901_C01_001E))) %>% 
  pivot_longer(cols = starts_with('S1901'), names_to = "name", values_to = 'num_household') %>% 
  mutate(name = str_remove(name, "E")) %>% 
  left_join(
    S1901_household %>% select(1:2) %>% 
      mutate(label = str_remove(label, "Estimate!!Households!!Total!!")),
            by = "name"
  )%>% 
  filter(label!="Estimate!!Households!!Total") %>% 
  select(-name) %>% 
  mutate(grp = case_when(
    label =='Less than $10,000'		~ 1,	
    label =='$10,000 to $14,999'	~ 2,	
    label =='$15,000 to $24,999'	~ 3,	
    label =='$25,000 to $34,999'	~ 4,	
    label =='$35,000 to $49,999'	~ 5,	
    label =='$50,000 to $74,999'	~ 6,	
    label =='$75,000 to $99,999'	~ 7,	
    label =='$100,000 to $149,999'	~ 8,		
    label =='$150,000 to $199,999'  ~ 9,
    label =='$200,000 or more'  ~ 9,
    TRUE ~ NA
  ), .after = label
  ) %>% 
  group_by(fipscode,statecode,countycode, grp) %>% 
  summarise(num_household = sum(num_household, na.rm = TRUE), .groups = "drop") 

acs_2022_v153_income_num_hshd
```




### disparity

```{r no cis }
v153_disparity_income <- purrr::reduce(
  list(acs_2022_v153_income_num_hshd, acs_2022_v153_income_num_den),
  dplyr::left_join,
  by = c('fipscode', 'statecode', 'countycode', 'grp')
) %>% 
  mutate(rawvalue = numerator/denominator) %>% 
  mutate(num_hh_wt = if_else(is.na(rawvalue), 0, num_household), .after = 'num_household') %>% 
  mutate(num_hh_th = if_else(rawvalue>0 & !is.na(rawvalue), num_household, 0)) %>% # set num_household=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode) %>% 
  mutate(hh_i = num_hh_wt / sum(num_hh_wt), .after = num_hh_wt) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_hh = sum(num_household), 
            num_hh_counted = sum(num_hh_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, hh_i, na.rm = TRUE),
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_hh_counted - 1)/num_hh_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(hh_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(hh_i * (rawvalue - weighted.mean(rawvalue, w = hh_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_hh_theil_counted = sum(num_hh_th, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, num_hh_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, num_hh_th),
            )

```


```{r with cis }
#note the rawvals are calcd above in the same step where the cis are calcd 
# therefore, rawvals are already available in acs_2022_v153_income_num_den_cis 
v153_disparity_income_cis <- purrr::reduce(
  list(acs_2022_v153_income_num_hshd, acs_2022_v153_income_num_den_cis),
  dplyr::left_join,
  by = c('fipscode', 'statecode', 'countycode', 'grp')
) %>% 
  mutate(num_hh_wt = if_else(is.na(rawvalue), 0, num_household), .after = 'num_household') %>% 
  mutate(num_hh_th = if_else(rawvalue>0 & !is.na(rawvalue), num_household, 0)) %>% # set num_household=0 if a tract has missing or 0 raw value, i.e, exclude a tract in weighting if it has missing or 0 raw value
  group_by(fipscode) %>% 
  mutate(hh_i = num_hh_wt / sum(num_hh_wt), .after = num_hh_wt) %>% 
  mutate(rawvalue_th = if_else(rawvalue ==0|is.nan(rawvalue), NA, rawvalue)) %>% 
  summarise(statecode = first(statecode), countycode = first(countycode),
            num_hh = sum(num_household), 
            num_hh_counted = sum(num_hh_wt),
            mean = mean(rawvalue, na.rm = TRUE),
            mean_wt = stats::weighted.mean(rawvalue, hh_i, na.rm = TRUE),
            sd_s = sd(rawvalue, na.rm = TRUE), # sample sd
            sd_p =  sqrt((num_hh_counted - 1)/num_hh_counted) * sd(rawvalue, na.rm = TRUE), # population SD
            bgv = sum(hh_i * (rawvalue - mean(rawvalue, na.rm = TRUE))^2, na.rm = TRUE),
            sd_bgv = sqrt(bgv),
            bgv_wt = sum(hh_i * (rawvalue - weighted.mean(rawvalue, w = hh_i, na.rm = TRUE))^2, 
                         na.rm = TRUE),
            sd_bgv_wt = sqrt(bgv_wt),
            num_hh_theil_counted = sum(num_hh_th, na.rm = TRUE),
            mean_th = mean(rawvalue_th, na.rm = TRUE),
            mean_th_wt = stats::weighted.mean(rawvalue_th, num_hh_th, na.rm = TRUE),
            theil = theil_wtd_gl(rawvalue),
            theil_wt = theil_wtd_gl(rawvalue, num_hh_th),
            )
```


# save outputs 
```{r}
save(v153_disparity_income, file = "calcd_data_how/income.RData")
save(v153_disparity_income_cis, file = "calcd_data_how/income_cis.RData")
save(v153_disparity_race, file = "calcd_data_how/race.RData")
save(v153_disparity_race_cis, file = "calcd_data_how/race_cis.RData")
save(v153_disparity_tract, file = "calcd_data_how/tract.RData")
save(v153_disparity_tract_cis, file = "calcd_data_how/tract_cis.RData")
```









# 3b. income levels:tracts

## owner, renter by income

```{r}
income_owner
```

```{r}
# tract level
acs_2022_B25118_tract <- get_acs_tract_data(year = yr, 
                                               variables = c(income_owner$name,income_renter$name))

acs_2022_B25118_tract
```

### owner

```{r}
# owner
acs_2022_B25118_tract_owner <- acs_2022_B25118_tract %>% 
  select(1:4, contains(income_owner$name)) %>% 
  select(1:4, ends_with("E")) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'numerator') %>% 
  mutate(name = str_remove(name, "E")) %>% 
  left_join(income_owner %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Owner occupied:!!")),
            by = "name"
  )
```

### renter
```{r}
# renter
acs_2022_B25118_tract_rent <- acs_2022_B25118_tract %>% 
  select(1:4, contains(income_renter$name)) %>% 
  select(1:4, ends_with("E")) %>% 
  pivot_longer(cols = starts_with('B25118'), names_to = "name", values_to = 'denominator') %>% 
  mutate(name = str_remove(name, "E")) %>% 
  left_join(income_renter %>% 
              select(1:2) %>% 
              mutate(label = str_remove(label, "Estimate!!Total:!!Renter occupied:!!")),
            by = "name"
  )

acs_2022_B25118_tract_rent
```

### numerator + denominator

```{r}
acs_2022_v153_income_num_den_tract <- acs_2022_B25118_tract_owner %>% 
  select(-name) %>% 
  left_join(acs_2022_B25118_tract_rent %>% select(-name),
            by = c('GEOID', 'fipscode', 'statecode', 'countycode', 'label')) %>% 
  select(1:4, label, numerator, denominator) %>% 
  # filter(is.na(numerator) | is.na(denominator)) %>% 
  mutate(denominator = numerator + denominator) %>% 
  mutate(grp = case_when(
    label =='Less than $5,000'		~ 1,	
    label =='$5,000 to $9,999'		~ 1,	
    label =='$10,000 to $14,999'	~ 2,	
    label =='$15,000 to $19,999'	~ 3,	
    label =='$20,000 to $24,999'	~ 3,	
    label =='$25,000 to $34,999'	~ 4,	
    label =='$35,000 to $49,999'	~ 5,	
    label =='$50,000 to $74,999'	~ 6,	
    label =='$75,000 to $99,999'	~ 7,	
    label =='$100,000 to $149,999'	~ 8,		
    label =='$150,000 or more'      ~ 9,
    TRUE ~ NA
  ), .after = label
  ) %>% 
  group_by(GEOID, fipscode,statecode,countycode, grp) %>% 
  summarise(numerator = sum(numerator), denominator = sum(denominator), .groups = "drop") 

acs_2022_v153_income_num_den_tract
```

## hoursefold: tract
```{r}
S1901_household <- v22_s %>% 
  filter(str_detect(name, "S1901")) %>% 
  slice(1:11)

S1901_household %>% select(1:2)
```


```{r}
acs_2022_S1901_tract <- get_acs_tract_data(year = yr, variables = c(S1901_household$name))

acs_2022_S1901_tract
```

# v153 R2024

```{r}
v153_r2025 <- get_acs_data(year = yr, variables = c('B25003_001', 'B25003_002'))

v153_r2025_2 <- v153_r2025 %>% 
  filter(countycode != "000") %>% 
  select(fipscode = 1, 2:3, starts_with("B25003") ) %>% 
  rename(numerator = B25003_002E, 
         denominator = B25003_001E) %>% 
  mutate(rawvalue = numerator/denominator,
         se_num_square = B25003_002M**2,
         se_denom_square = B25003_001M**2) %>% 
  mutate(se_num = (sqrt(se_num_square))*1.96/1.645,
         se_denom = (sqrt(se_denom_square))*1.96/1.645) %>% 
  mutate(se_test = se_num^2 - (rawvalue^2 * se_denom^2)) %>% 
  mutate(se_pct = case_when(
    se_test>=0 ~ sqrt(se_num**2 - (rawvalue^2 * se_denom^2))/denominator,
    se_test< 0 ~ sqrt(se_num**2 + (rawvalue^2 * se_denom^2))/denominator
  )) %>% 
  mutate(cilow = rawvalue - se_pct, 
         cihigh = rawvalue + se_pct,
         sourceflag = NA_real_)%>% 
  select(c(1:3, denominator,numerator,rawvalue,cilow,cihigh))

v153_r2025_2 
```

```{r}
v153_r2025_2 %>% 
  filter(countycode!= "000", statecode == "55") %>% 
  ggplot(aes(fipscode, rawvalue)) +
  geom_line(group = 1) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "v153 - WI")

```



```{r}
# mean_wt
plotly::ggplotly(
  v153_disparity_TractRaceTandR %>% 
  filter(str_detect(fipscode, "^55")) %>% 
  select(1, mean_wt_Tract, mean_wt_Race, mean_wt_RandT) %>% 
  pivot_longer(cols = c(2:4), names_to = "var", values_to = "mean_wt") %>% 
  ggplot(aes(fipscode, mean_wt, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "weighted mean - WI")
)
```

## compare v153 with weighted means
```{r}
ggplotly(
  v153_r2025_2 %>% 
  filter(countycode!= "000", statecode == "55") %>% 
  select(fipscode, v153 = rawvalue) %>% 
  left_join(
    v153_disparity_TractRaceTandR %>% 
      filter(str_detect(fipscode, "^55")) %>% 
      select(1, mean_wt_Tract, mean_wt_Race, mean_wt_RandT),
    by = "fipscode"
  ) %>% 
  pivot_longer(cols = c(2:5), names_to = "var", values_to = "mean_wt") %>% 
  ggplot(aes(fipscode, mean_wt, group = var, color = var)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "WI")
)

```

# -------------
